% =============================================================================
%  SŌKYŪ-Ō SŌBA ZENSHŪ — Complete Works of Old Man Sōkyū on Markets
%  First English Translation
%
%  Source text by: 本間宗久 (Honma Sōkyū)
%  Annotated by: 早坂豐蔵 (Hayasaka Hōzō)
%  Published: 信義堂書店 (Shin'gidō Shoten), Tokyo, Meiji 43 (1910)
%
%  Edited and produced by Zach Booth
%  Translated with AI assistance (Claude, Anthropic)
%
%  Compiled with XeLaTeX
% =============================================================================

\documentclass[10pt, openany]{book}

\usepackage[
  papersize={6in,9in},
  margin=0.75in,
  inner=0.85in,
  outer=0.65in,
  top=0.8in,
  bottom=0.8in
]{geometry}

\usepackage{fontspec}
\setmainfont{Palatino}[
  BoldFont={Palatino Bold},
  ItalicFont={Palatino Italic},
  BoldItalicFont={Palatino Bold Italic},
]
\setsansfont{Helvetica Neue}
\newfontfamily\japanesefont{Hiragino Mincho ProN}[
  AutoFakeBold=2.5,
]
\newcommand{\jp}[1]{{\japanesefont #1}}

\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{titlesec}
\usepackage{fancyhdr}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{microtype}
\emergencystretch=1em
\usepackage{hyperref}
\usepackage{float}
\usepackage{setspace}
\usepackage{lettrine}
\usepackage{parskip}
\usepackage{needspace}
\usepackage{mdframed}

\definecolor{chaptercolor}{HTML}{8B0000}
\definecolor{sectioncolor}{HTML}{333333}

\titleformat{\chapter}[display]
  {\normalfont\Large\bfseries\color{chaptercolor}}
  {\chaptertitlename\ \thechapter}{12pt}{\LARGE}
\titlespacing*{\chapter}{0pt}{-20pt}{20pt}

\titleformat{\section}
  {\needspace{8\baselineskip}\normalfont\large\bfseries\color{sectioncolor}}
  {\thesection}{1em}{}

\titleformat{\subsection}
  {\needspace{3\baselineskip}\normalfont\normalsize\bfseries\itshape\color{sectioncolor}}
  {\thesubsection}{1em}{}

\widowpenalties 3 10000 10000 150
\clubpenalties 3 10000 10000 150

\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE]{\small\itshape S\=oky\=u-\=o S\=oba Zensh\=u}
\fancyhead[RO]{\small\itshape\leftmark}
\fancyfoot[CE,CO]{\small\thepage}
\renewcommand{\headrulewidth}{0.4pt}

% Translator's note footnote
\newcommand{\tn}[1]{\footnote{\textit{Translator's note:} #1}}

% Japanese term with romanization gloss
\newcommand{\jt}[2]{\jp{#1} (\textit{#2})}

% Editorial note box
\definecolor{editorbg}{HTML}{F5F2EB}
\definecolor{editorbar}{HTML}{BBBBBB}
\newmdenv[
  topline=false,
  bottomline=false,
  rightline=false,
  linewidth=2pt,
  linecolor=editorbar,
  backgroundcolor=editorbg,
  innerleftmargin=10pt,
  innerrightmargin=10pt,
  innertopmargin=6pt,
  innerbottommargin=6pt,
  skipabove=8pt,
  skipbelow=8pt,
]{editorbox}
% Modern editorial notes (by the present editor/translator)
\newenvironment{editorialnote}{%
  \begin{editorbox}\small\textit{Editor's note:}\enspace}{\end{editorbox}}

% Hayasaka's original annotations (校註)
\definecolor{hayasakabg}{HTML}{F0EDE5}
\definecolor{hayasakabar}{HTML}{999999}
\newmdenv[
  topline=false,
  bottomline=false,
  rightline=false,
  linewidth=2pt,
  linecolor=hayasakabar,
  backgroundcolor=hayasakabg,
  innerleftmargin=10pt,
  innerrightmargin=10pt,
  innertopmargin=6pt,
  innerbottommargin=6pt,
  skipabove=8pt,
  skipbelow=8pt,
]{hayasakabox}
\newenvironment{hayasakanote}{%
  \begin{hayasakabox}\small\textit{Hayasaka's annotation (\jp{校註}):}\enspace}{\end{hayasakabox}}

\renewcommand{\chaptermark}[1]{\markboth{#1}{}}

\hypersetup{
  colorlinks=true,
  linkcolor=chaptercolor,
  urlcolor=blue!60!black,
  pdftitle={Sōkyū-ō Sōba Zenshū: Complete Works of Old Man Sōkyū on Markets},
  pdfauthor={Edited by Zach Booth; translated with AI assistance (Claude, Anthropic)},
}

\pdfstringdefDisableCommands{\def\jp#1{}}
\pdfstringdefDisableCommands{\def\jt#1#2{#2}}

% =============================================================================
\begin{document}

% --- Half-title ---
\thispagestyle{empty}
\begin{center}
\vspace*{2.5in}
{\Large\itshape S\=oky\=u-\=o S\=oba Zensh\=u}\\[0.5em]
{\large Complete Works of Old Man S\=oky\=u on Markets}
\vfill
\end{center}
\newpage\thispagestyle{empty}\mbox{}\newpage

% --- Title page ---
\thispagestyle{empty}
\begin{center}
\vspace*{1in}
{\huge\bfseries\color{chaptercolor} \jp{宗久翁相場全集}}\\[1em]
{\LARGE\bfseries S\=oky\=u-\=o S\=oba Zensh\=u}\\[0.7em]
{\Large\itshape Complete Works of Old Man\\S\=oky\=u on Markets}\\[2em]
\rule{3in}{0.5pt}\\[1.5em]
{\large\itshape The Collected Wisdom of\\Honma S\=oky\=u on Rice Trading}\\[0.5em]
{\normalsize from the Edo Period of Japan}\\[2.5em]
{\normalsize Posthumous manuscripts of}\\[0.3em]
{\large \jp{本間古作}}\\
{\normalsize Honma Kosaku (S\=oky\=u)}\\[0.5em]
{\normalsize of Sakata, Dewa Province}\\[1.5em]
{\normalsize Annotated and compiled by}\\[0.3em]
{\large \jp{早坂豐蔵}}\\
{\normalsize Hayasaka H\=oz\=o}\\[2em]
\rule{3in}{0.5pt}\\[1.5em]
{\normalsize Edited and produced by Zach Booth}\\[0.3em]
{\small Translated with AI assistance}\\[3em]
\end{center}
\vfill
\newpage

% --- Copyright / attribution ---
\thispagestyle{empty}
~\vfill

\begin{small}
\noindent\textbf{Source Text}\\[0.3em]
\noindent\jp{宗久翁相場全集} (\textit{S\=oky\=u-\=o S\=oba Zensh\=u})\\
Published: Meiji 43 (1910) by \jp{信義堂書店} (Shin'gid\=o Shoten), Tokyo\\
Author: \jp{本間宗久} (Honma S\=oky\=u, 1724--1803)\\
Editor: \jp{早坂豐蔵} (Hayasaka H\=oz\=o)\\
Source scans: Digital archive, catalog number 804042\_0001\\[1em]

\noindent\textbf{About This Edition}\\[0.3em]
\noindent This is the first English translation of the \textit{S\=oky\=u-\=o
S\=oba Zensh\=u}. The translation was produced in two stages: first a
character-by-character transcription of the Japanese woodblock print
with confidence annotations, then an English translation from that
transcription. Both the transcription and translation files are
maintained alongside the original scans for full auditability.\\[0.5em]
\noindent This book is the second volume in the series, following
\textit{Sanen Kinsen Hiroku} (Three Monkeys Gold Spring Secret Record).\\[1em]

\noindent English translation copyright \textcopyright{} 2026 Zach Booth.
All rights reserved.
\end{small}
\newpage

% =============================================================================
\thispagestyle{empty}
\tableofcontents
\newpage

% =============================================================================
\chapter*{Translator's Introduction}
\addcontentsline{toc}{chapter}{Translator's Introduction}
\markboth{Translator's Introduction}{}

The \textit{S\=oky\=u-\=o S\=oba Zensh\=u} (\jp{宗久翁相場全集}), or
``Complete Works of Old Man S\=oky\=u on Markets,'' is a collection of
trading wisdom attributed to Honma S\=oky\=u (1724--1803), widely regarded
as the most legendary rice trader in Japanese history.\tn{Honma S\=oky\=u,
also known by his given name Shingozaemon, traded rice at the Sakata,
\=Osaka D\=ojima, and Edo exchanges during the mid-Edo period. The Honma
family of Sakata amassed one of the greatest fortunes in Japanese
commercial history.}

The text was compiled posthumously from S\=oky\=u's scattered manuscripts
by the Meiji-era editor Hayasaka H\=oz\=o, who collected, annotated, and
published them as a two-volume work in 1910. Until that time, the
writings had circulated only in hand-copied form, with the inevitable
accumulation of errors and omissions.

\section*{Structure of the Text}

The work is divided into two volumes:

\begin{itemize}
\item \textbf{Upper Volume} (\jp{上巻}, \textit{j\=okan}): 73 chapters
covering the fundamental principles of market trading---the psychology
of crowds, seasonal patterns, position management, risk control, and
the cultivation of the trader's mind.

\item \textbf{Lower Volume} (\jp{下巻}, \textit{gekan}): 66 chapters
covering practical techniques---calendar-based timing, specific price
benchmarks, regional market relationships, and empirical data from the
An'ei era (1772--1781).
\end{itemize}

Together the 139 chapters form a comprehensive manual of market
wisdom that remains remarkably relevant to modern trading.

\section*{Key Concepts}

Several principles recur throughout the text:

\begin{description}[style=nextline]
\item[The overall trend (\jp{大勢}, \textit{taisei})]
Never trade against the prevailing direction. Small fluctuations
within a larger trend are noise; the trend itself is signal.

\item[The three pivotal months (\jp{三位の月}, \textit{sanmi no tsuki})]
January, May, and September mark the turning points of the market
cycle. Major trend changes cluster around these months.

\item[Three-tenths profit (\jp{三つ十分の極し})]
Take three-tenths of the available profit and consider it complete.
Greed for the full ten-tenths leads to loss.

\item[The three-test rule]
Neither bottoms nor ceilings are confirmed until tested three times.
Only after the third test should one commit to a position.

\item[Patience (\jp{辛抱}, \textit{shinb\=o})]
The market rewards those who wait. Great opportunities come only
once in three to five years.
\end{description}

\section*{A Note on the Text}

Within each chapter, subsections marked \textit{Seed} (\jp{種},
\textit{tane}) contain supplementary teachings---kernels of wisdom
that elaborate on, contrast with, or distill the chapter's main
point. These appear to be part of S\=oky\=u's original text. Numbered
sections marked \textbf{(1)}, \textbf{(2)}, etc.\ denote sequential
points within a teaching.

Two types of annotation appear in shaded boxes. Passages marked
\textit{Hayasaka's annotation} (\jp{校註}) are notes by the 1910
editor Hayasaka H\=oz\=o, who compiled and annotated the original
manuscripts. These are part of the source text. Passages marked
\textit{Editor's note} are annotations by the present editor and
translators, providing historical context or structural explanation.
Footnotes marked \textit{Translator's note} provide definitions and
comparative observations.

\vfill
\begin{flushright}
\textit{---Zach Booth}\\
\textit{March 2026}
\end{flushright}
\newpage

% =============================================================================
\part{Upper Volume \jp{（上巻）}}

% =============================================================================
% UPPER VOLUME CHAPTER 1
% =============================================================================
\chapter{Grain Is the Great Foundation of the Nation}
\markboth{Chapter 1: Grain Is the Great Foundation}{}

\begin{center}
\jp{第一章　水穀は國の大本なるの事}
\end{center}

\noindent Of all the great treasures under heaven, nothing
compares to rice and grain. Of all the things that sustain human life,
nothing surpasses it. Therefore, anyone who wishes to master the way of
the market must first understand the fundamental importance of grain.

One must grasp this principle, understand the essential purpose of the
market, and only then study the methods of trading. Rice is the great
foundation of the nation; the people of the realm depend upon it for
their livelihood. The price of one \textit{koku} (\jp{石}) of rice rises
and falls according to the quality of the harvest (\jt{豊凶}{h\=oky\=o}).
In years of abundance, prices are low; in years of famine, prices are
high. This is the natural order of things. Yet the market, beyond this
natural order, rises and falls greatly according to the inclination of
popular sentiment (\jt{人氣}{ninki}). Through people's speculation,
prices rise still higher when already high, and fall still lower when
already low. This is what makes a market a market.

\bigskip

\textbf{(1)} Since the price of rice is thus determined by a combination of the
natural principles of heaven and earth and the inclinations of popular
sentiment, anyone who intends to trade must first understand these two
principles.

In undertaking any trading, one must first assess the overall trend
(\jt{大勢}{taisei}) of the realm. Is it a good harvest, or a poor one?
What is the condition of the economy? What are the circumstances in the
various provinces? One must observe all of these carefully and only then
buy or sell. To trade without knowing the overall trend, relying only on
the prices before one's eyes, is like a blind man walking a road in the
dark of night.

\bigskip

\textbf{(2)} The market is the human heart. To see into people's hearts is the
essential point of the way of the market. When everyone thinks of buying,
the price of rice will certainly rise. When everyone thinks of selling,
the price of rice will certainly fall. And yet, when everyone is buying,
the ceiling is near; when everyone is selling, the bottom is near.
Therefore, to go against the crowd's heart is the secret of the market.

\begin{hayasakanote}
This chapter sets forth the fundamental principles of the market.
The reader should savor its meaning well and apply it to the chapters
that follow.
\end{hayasakanote}

The way of the market is like water flowing to the lowest point. Water
flows from high to low and naturally levels itself. The market is the
same. When prices are high, they naturally fall; when prices are low,
they naturally rise. He who understands this principle may be said to
have grasped the way of the market. Yet people of the world do not
understand this principle---they go on buying when prices are high and
selling when prices are low. This is a great error.

In the market, there is a great upheaval within every three-year period.
Taking three years as one cycle, there will be changes of harvest quality
within that span. Know these changes, buy in abundant years when prices
are low, and sell in famine years when prices are high. This is the
great way (\jt{大道}{taid\=o}) of the market. Do not forfeit great profits
by grasping at small ones.

Furthermore, in market trading, one must not listen to what others say.
Think for yourself, judge for yourself, and decide for yourself. To trade
by following what others say will certainly lead to failure.

The way of the market is deep and far-reaching. It cannot be mastered in
a single morning or evening. One must cultivate it over years and months,
studying day and night, before one can attain its subtlety.

% =============================================================================
% UPPER VOLUME CHAPTER 2
% =============================================================================
\chapter{Do Not Buy Declining Grain at High Prices}
\markboth{Chapter 2: Do Not Buy Declining Grain at High Prices}{}

\begin{center}
\jp{第二章　下穀は高値にて買ふべからずの事}
\end{center}

\noindent In trading, one must not buy declining grain
(\jt{下穀}{shitakoku}) at high prices. ``Declining grain'' refers to
rice whose price is falling. When the price of rice has dropped,
everyone is frightened and wants to sell. To buy at such a time is like
catching a falling blade with one's bare hands.\tn{The Japanese is
\jp{落つる刃を手にて受くるが如し}---literally ``like receiving a
falling blade with one's hands.'' This metaphor, which S\=oky\=u uses
three times throughout the text (Teachings 2, 68, and Lower~17),
predates by well over a century the modern Wall Street adage ``don't
catch a falling knife.'' Whether the Western saying derives from the
Japanese trading tradition or arose independently is unknown, but
S\=oky\=u's use of it in the late eighteenth century is among the
earliest known applications of this image to financial markets.}

\bigskip

\textbf{(1)} To ``buy declining grain at high prices'' means to see a slight
rally in the middle of a falling market, think that the bottom has been
reached, and rush in to buy. This is a great error.

In a falling market, even if prices rise slightly, they will immediately
fall again. Therefore, to see a small rally and think the bottom has been
reached is a great misunderstanding. Before concluding that a bottom has
been reached, one should wait through three tests, and only then make a
determination.

The price hits a bottom and rises, then falls again without breaking the
previous bottom, and rises; then falls again without breaking the
previous bottom, and rises. When the bottom has been tested three times
in this way without breaking, then for the first time one may know the
bottom is firm. This is the time to buy.

\begin{hayasakanote}
This chapter is related to what is known as the method of the ``three
rivers'' (\jt{三川}{sankawa}), a foundational pattern in Japanese
candlestick analysis.
\end{hayasakanote}

Therefore, declining grain should be bought only after the firmness of
the bottom has been confirmed. One must not buy in haste at high prices.
If one buys in haste, one will certainly incur losses. Patience
(\jt{辛抱}{shinb\=o}) is the first principle of the way of the market.

% =============================================================================
% UPPER VOLUME CHAPTER 3
% =============================================================================
\chapter{Do Not Plunge Headlong into Greatly Risen Grain}
\markboth{Chapter 3: Do Not Plunge into Greatly Risen Grain}{}

\begin{center}
\jp{第三章　引は大穀の説大中へ飛込む心の事}
\end{center}

\noindent When one sees rice that has risen greatly in the
market and thinks ``I too shall buy,'' plunging in to purchase
it---this is a great error. That which has risen greatly will certainly
fall greatly.

\bigskip

\textbf{(1)} Everyone, seeing that prices have risen greatly, thinks they will
rise still further and plunges in to buy. This is foolishness. What has
risen will fall, and what has fallen will rise. This is the natural
principle of the market.

To see rice that has greatly risen and think in one's heart, ``it will
surely rise further''---this is human nature. And yet, this is the
market's great enemy. After a great rise, there will certainly be a
fall; therefore this is a time for selling, not a time for buying.

Yet people of the world, not knowing this principle, buy when prices
have risen greatly and sell when prices have fallen greatly. This is
the reason they lose in the market. This principle must not be forgotten.

% =============================================================================
% UPPER VOLUME CHAPTER 4
% =============================================================================
\chapter{Seasonal Patterns and Crowd Psychology}
\markboth{Seasonal Patterns and Crowd Psychology}{}

\section*{Teaching 4: The Spirit of Riding the Crowd's Momentum Upward}
\addcontentsline{toc}{section}{Teaching 4: The Spirit of Riding the Crowd's Momentum Upward}

\begin{center}
\jp{第四章　人心に上りの勢津中へ大人なる心の事}
\end{center}

\noindent In the market, it is essential to understand where
the hearts of the crowd are heading. When the crowd's heart all thinks
prices will rise, the market has already reached near the ceiling
(\jt{天井}{tenj\=o}). When the crowd's heart all thinks prices will
fall, the market has already reached near the bottom (\jt{底}{soko}).
He who understands this principle and goes against the crowd's heart
is the one who wins in the market.

When the price of rice has risen greatly and everyone, thinking it will
rise still further, scrambles to buy---the market will certainly fall
before long. He who at this time makes his heart bold and tries to buy
will incur great losses.

\bigskip

\textit{Seed:}\tn{Throughout the text, sections marked \textit{Seed}
(\jp{種}, \textit{tane}) denote supplementary teachings within a
chapter. The character \jp{種} literally means ``seed''---a kernel of
wisdom planted for the reader to cultivate. These sections sometimes
elaborate on the main teaching, sometimes offer a contrasting
perspective, and sometimes state a rule in its most condensed form.
They function somewhat like marginalia or glosses in Western
manuscripts, and appear to be part of the original text rather than
later additions by the editor.} In a rising market, to ride the crowd's momentum and buy
is the same as buying the ceiling. He who buys the ceiling will
certainly suffer great losses. Therefore, when everyone is buying in a
rising market, that is precisely the time to sell. This principle is the
secret of the way of the market; engrave it deeply upon your heart.

% =============================================================================
% UPPER VOLUME CHAPTER 5
% =============================================================================
\section*{Teaching 5: The Winter-to-New-Year Rice Market}
\addcontentsline{toc}{section}{Teaching 5: The Winter-to-New-Year Rice Market}

\begin{center}
\jp{第五章　冬より正月迄相場合米の事}
\end{center}

\noindent From winter through the New Year, the rice market
generally rises. The reason is that winter is the season when demand for
rice increases, and moreover the consumption of rice and grain is heavy
due to New Year's celebrations and the like.

\bigskip

\textbf{(1)} Therefore, one should buy and stock up from around October, and sell
when the New Year arrives. This is the normal way of the winter-to-New-Year
market.

However, depending on the year, there are exceptions to this rule. In
the year following a famine year, prices may still be high even into
spring. In the year following an abundant year, prices may not rise even
in winter. Knowing these exceptions is also essential.

\begin{hayasakanote}
``New Year'' here refers to the period from December 28th through
January 15th.
\end{hayasakanote}

Thus, understand that the winter-to-New-Year market normally rises, and
stock up from around October. However, as noted previously, there are
exceptions depending on the year, so one should assess the overall trend
(\jp{大勢}, \textit{taisei}) before committing.

% =============================================================================
% UPPER VOLUME CHAPTER 6
% =============================================================================
\section*{Teaching 6: Markets that Fall Sharply}
\addcontentsline{toc}{section}{Teaching 6: Markets that Fall Sharply}

\begin{center}
\jp{第六章　急に下げ値にても相場の事}
\end{center}

\noindent When the market has fallen sharply, everyone is
frightened and sells. Yet that which has fallen sharply may also rise
sharply again. After a crash comes a rebound. He who understands this
principle should buy without fear at the time of a crash.

\bigskip

\textbf{(1)} However, among sharp declines there are both genuine declines and
temporary declines. A genuine decline is driven by the overall trend of
the realm, and the rebound is small. A temporary decline is driven by
the wavering of crowd sentiment (\jp{人氣}, \textit{ninki}), and prices
return quickly. To distinguish between these two is essential.

% =============================================================================
% UPPER VOLUME CHAPTER 7
% =============================================================================
\section*{Teaching 7: The Misfortune of a Winter-to-New-Year Ceiling Followed by Falling Prices}
\addcontentsline{toc}{section}{Teaching 7: The Misfortune of a Winter-to-New-Year Ceiling Followed by Falling Prices}

\begin{center}
\jp{第七章　冬より正月迄天井落値の不首尾の事}
\end{center}

\noindent It sometimes happens that the market reaches its
ceiling from winter through the New Year, and thereafter begins to fall.
Such a case is called \jt{不首尾}{fushubi} (failure, misfortune). That
is, positions bought during winter fail to rise after the New Year has
passed, and instead begin to fall.

The causes of this failure are various, but it most often occurs in
years of good harvests. In years of good harvests, rice continues to
come to market abundantly even in winter, and after the New Year passes,
even more comes to market; therefore the price hits its ceiling and
falls.

Therefore, in the winter market, one must carefully assess the harvest
quality of the year, then commit. In an abundant year, the ceiling may
come early, and the trade may end in failure. This principle must not be
forgotten.

\bigskip

Any trade that ends in failure (\jp{不首尾}, \textit{fushubi}) should be
cut quickly. If one clings to it and holds on, losses only grow larger.

% =============================================================================
% UPPER VOLUME CHAPTER 8
% =============================================================================
\section*{Teaching 8: The Price Decline of July through October}
\addcontentsline{toc}{section}{Teaching 8: The Price Decline of July through October}

\begin{center}
\jp{第八章　七、八、九、十月茶落値下げの事}
\end{center}

\noindent From July through October, rice prices generally
fall. This is because it is the season when new-crop rice comes to
market, and the supply of grain is abundant. August and September in
particular are when early-harvest rice first comes to market, and
prices decline. To buy during this season is a very dangerous thing.

\bigskip

\textbf{(1)} During the months of July, August, September, and October, one
should not buy rice. Since it is the time when new-crop rice comes to
market, it is the natural order for prices to fall. However, in years of
famine, this rule does not apply. In famine years, the supply of
new-crop rice is small, and prices do not fall---indeed they may rise.

% =============================================================================
% UPPER VOLUME CHAPTER 9
% =============================================================================
\chapter{Being Prepared for Failure}
\markboth{Chapter 9: Being Prepared for Failure}{}

\begin{center}
\jp{第九章　不首尾覺悟の事}
\end{center}

\noindent In the market, it is essential to be prepared
(\jt{覺悟}{kakugo}) for failure (\jt{不首尾}{fushubi}). The market moves
according to the time of heaven, the advantage of the ground, and the
harmony of men---and there is much that lies beyond human power.
Therefore, no matter how much wisdom one exhausts or how much ingenuity
one devises, it is impossible to win at all times.

When a trade fails, one should quickly cut one's losses. To stop with a
small loss is the way to prevent a great loss.

\bigskip

\textbf{(1)} When one has lost in the market, do not lose heart; calmly
consider the causes. The same applies when one has won. Consider why one
won. He who knows the causes of victory and defeat will certainly have
an advantage in the next contest.

Furthermore, when a trade fails, one should rest for a while. During a
losing streak, the mind becomes disordered and one often misjudges.
Rest, calm the mind, and only then begin anew.

\bigskip

\textit{Seed:} The market is patience (\jp{辛抱}, \textit{shinb\=o}).
The patient will certainly win; the impatient will certainly lose. This
one thing---engrave it well upon your heart.

\bigskip

Victory and defeat in the market depend largely upon the mandate of
heaven. One should approach the market with the mindset of doing
everything humanly possible and then awaiting heaven's decree. He who
rushes to win will certainly lose; he who fears losing will also lose.
To approach the market with an ordinary mind---this is the ultimate
secret of the market.

Furthermore, in the market, capital management is essential. One must
not trade with one's entire fortune. To trade with one-tenth is
considered good. Even if one meets with failure, one can recover and
start again.

\bigskip

Even in a small trade, one must not bet large capital. \textit{Haru}
(\jp{張る}) means to wager---to commit large sums. To commit large sums
for the sake of a small profit is like using a golden fishing rod to
catch a sea bream. The profit is small and the danger is great.

\bigskip

\textit{Seed:} The first principle of trading should be: above all, do
not incur losses. To put gaining profit first and preventing loss second
is to have one's priorities inverted. First prevent loss, then seek
profit. This is the true way (\jt{正道}{sh\=od\=o}) of the market.

A farmer does not stop plowing his fields even after a small poor
harvest. Within twelve or thirteen years, there will be one great bumper
crop. The market is the same. Within three to five years, there will
certainly be one great opportunity. Do not let this opportunity
escape---seize your profit.

\bigskip

\textit{Seed:} Within a twelve- or thirteen-month period, major
upheavals are normal. Between January and March, and between July and
September---these two periods are when major upheavals most often occur.
Pay attention during these periods, watch for the moment, and act.

In the market, one must know the time of heaven, discern the harmony of
men, and understand the advantage of terrain---only then can one win.
``The time of heaven'' means the changes of harvest quality. ``The
advantage of terrain'' means the price differences between regions. ``The
harmony of men'' means the inclinations of popular sentiment. He who
combines knowledge of all three will face a hundred battles without
danger.

This is the ultimate meaning of the chapter on being prepared for
failure. Engrave it well upon your heart.

\bigskip

Furthermore, in trading, do not listen to the opinions of many people.
If you listen to the opinions of many, your mind will be confused and
you will be unable to make decisions. Trust your own judgment and act
decisively.

The market is a solitary path. He who walks with the crowd cannot win
in the market. Only he who walks the path the crowd does not walk will
gain profit.

Furthermore, as a principle of trading, it is precisely when one has
won that one should exercise caution. To bet heavily on the heels of a
victory is the foundation of defeat. When you have won, hold back a
little; when you have lost, rest. This is the ultimate secret of the
market.

\bigskip

Additionally, the following principles of trading are here recorded:

\begin{enumerate}
\item Before entering the market, determine the conditions for retreat.
\item Exercise restraint in pyramiding profits.
\item Rid yourself of the desire to capture a huge trade.
\item Cut losses quickly. Take profits slowly.
\item When your forecast proves wrong, correct it immediately.
\end{enumerate}

He who observes these five rules will never suffer a great failure in
the market.

The market is not a matter of intellect. It is a matter of the heart.
He whose heart is right will gain profit; he whose heart is crooked will
incur loss. If one understands this principle, one may be said to have
passed the halfway point of the way of the market.

\bigskip

\textit{Seed:} To trade, one must first know the art of war
(\jt{軍法}{gunp\=o}). The market is a battle. ``He who knows the enemy
and knows himself will face a hundred battles without danger.'' The
enemy means the movement of the market; the self means one's own
capital and strength of heart.

One must not fight battles that exceed one's strength. To wage a great
contest with small capital will certainly end in defeat. Fight battles
suited to your strength, and do not attempt the impossible.

In the market, it is better to accumulate many small victories than to
win one great victory. He who tries to win big all at once will lose big
all at once. Accumulate small profits to build great profit. This is the
true way of the market.

\bigskip

Thus the chapter on being prepared for failure sets forth the fundamental
principles of the trading mindset.

In the market, to win is heaven's decree; to lose is also heaven's
decree. Yet to leave matters to heaven's decree without exhausting human
effort is not the way of the market. Study the teachings of this book
thoroughly, accumulate practice and training, and only then approach the
market. This is both the beginning and the end of the way of the market.

This one chapter---let all who aspire to the market read it with care.

And moreover, even if one watches the market for eight or nine months
and takes a profit, one should still discern where the momentum has
changed.

% =============================================================================
% UPPER VOLUME CHAPTER 10
% =============================================================================
\chapter{Trend, Position, and Timing}
\markboth{Trend, Position, and Timing}{}

\section*{Teaching 10: The Market's Rises and Falls Following the Overall Trend}
\addcontentsline{toc}{section}{Teaching 10: The Market's Rises and Falls Following the Overall Trend}

\begin{center}
\jp{第十章　現場は上下大勢の事}
\end{center}

\noindent The rises and falls of the market depend on the
overall trend (\jp{大勢}, \textit{taisei}). The overall trend is the
combined picture of the economy of the realm, the harvest quality of
rice, the conditions in the various provinces, and so forth. One must
not be captivated by small fluctuations, but should assess the overall
trend and trade accordingly.

\bigskip

\textbf{(1)} When the overall trend is upward, prices will still rise even if
they fall slightly. When the overall trend is downward, prices will
still fall even if they rise slightly. He who misjudges the overall
trend cannot win in the market.

Do not let two or three days' fluctuations move your heart. Look at the
trend of one, two, three months and trade accordingly. To chase small
profits and lose sight of the overall trend is the constant error of the
market trader.

% =============================================================================
% UPPER VOLUME CHAPTER 11
% =============================================================================
\section*{Teaching 11: Averaging Down When Prices Move Against You}
\addcontentsline{toc}{section}{Teaching 11: Averaging Down When Prices Move Against You}

\begin{center}
\jp{第十一章　行相違平割段の事}
\end{center}

\noindent In the market, it sometimes happens that the price
of rice one has bought falls contrary to one's expectations. At such
times, one may buy more at the lower price to bring down the average.
This is called \jt{行相違平割}{yukis\=oi hirawari} (averaging down
against an adverse move).

For example, if one bought rice at thirty-five \textit{monme} (\jp{匁})
and the price fell to thirty-three \textit{monme}, one buys more at
thirty-three \textit{monme}. The average then becomes thirty-four
\textit{monme}, and one gains a profit if the price rises above
thirty-four \textit{monme}.

However, this method should be used with caution. If one uses this
method during a period when the overall trend is downward, losses will
only grow larger. Use this method only when the bottom is firm.

\begin{hayasakanote}
This is what is called today the method of ``averaging down''
(\jt{難平均}{nanpinkin}). In the original text from the Bunka era,
it reads ``to level out an adverse move.''
\end{hayasakanote}

In a consolidation, keep your mind calm and set a limit of six months.
If the price has not risen within that time, cut your losses and exit.
Do not wait too long and let losses grow.

\bigskip

\textit{Seed:} When prices have moved against you, first assess the
overall trend. When the trend is upward, average down and wait. When the
trend is downward, cut losses quickly and exit. To distinguish between
these two situations is essential.

Furthermore, do not average down in large increments all at once. Buy
more in small amounts. If you buy a large amount all at once, you will
have no reserve strength to buy more if prices fall still further.

% =============================================================================
% UPPER VOLUME CHAPTER 12
% =============================================================================
\section*{Teaching 12: The Three Pivotal Months and Extending Momentum}
\addcontentsline{toc}{section}{Teaching 12: The Three Pivotal Months and Extending Momentum}

\begin{center}
\jp{第十二章　三位青月延勢の事}
\end{center}

\noindent The ``three pivotal months'' (\jt{三位の月}{sanmi
no tsuki}) are January, May, and September. These three months are the
times when the market moves greatly and the overall trend undergoes a
shift. \textit{Ensei} (\jp{延勢}, extending momentum) means the
extension of a trend---when upward momentum or downward momentum
continues without stopping.

\bigskip

\textbf{(1)} At these three pivotal months, one should assess the overall trend
and trade accordingly. January is influenced by the momentum carried
over from winter; May is influenced by the momentum of spring; and
September is influenced by the momentum of summer.

\bigskip

\textit{Seed:} To know the three pivotal months and determine whether
extending momentum is present or absent---this is the essential point of
the way of the market. If extending momentum is present, ride that
momentum and take your profit. If extending momentum is absent, keep
your hands out of the market.

\bigskip

Furthermore, the duration of extending momentum should be reckoned in
three-month cycles. If the direction of momentum has not been
established within three months, wait for the next pivotal month.

January is the time for buying. Rice that bottomed out during winter
often begins to rise from January. May is the time for the ceiling. Rice
that rose during spring often hits its ceiling around May. September is
the time for the bottom. Rice that fell during summer often hits its
bottom around September.

\bigskip

\textit{Seed:} Since January, May, and September are the turning points
of the market, pay attention during these seasons and assess the overall
trend. To know the three pivotal months is a matter of great importance
in the way of the market.

% =============================================================================
% UPPER VOLUME CHAPTER 13
% =============================================================================
\section*{Teaching 13: The Winter Ceiling and June Decline}
\addcontentsline{toc}{section}{Teaching 13: The Winter Ceiling and June Decline}

\begin{center}
\jp{第十三章　冬相天井六月下げの事}
\end{center}

\noindent That the market reaches its ceiling
(\jp{天井}, \textit{tenj\=o}) in winter and declines by June is the
normal course of the market. In winter, demand for rice is high and
inventories are low; therefore prices rise. However, by May and June,
new wheat begins to come to market, and the prospects for early-harvest
rice become clear as well---and so the price of rice falls.

\bigskip

\textbf{(1)} Knowing this principle, sell at the winter ceiling and buy at the
June bottom. This is the great way of the market. However, in some years
this rule does not hold. In famine years, prices may continue to rise
beyond the winter ceiling. Know these exceptions as well.

Furthermore, the winter ceiling is most often reached in December or
January. The June bottom is most often reached in late May or early
June. Know these seasons and set the timing of your trades accordingly.

% =============================================================================
% UPPER VOLUME CHAPTER 14
% =============================================================================
\section*{Teaching 14: Entering in Two Parts and Taking Three-Tenths as Full Profit}
\addcontentsline{toc}{section}{Teaching 14: Entering in Two Parts and Taking Three-Tenths as Full Profit}

\begin{center}
\jp{第十四章　二つ仕掛三つ十分の極しの事}
\end{center}

\noindent When entering a position (\jt{仕掛}{shikake}),
one should divide it into two parts. To enter with one's full strength
all at once is a great error. First enter with one of the two parts,
observe the market's movement, and only then enter with the remaining
part.

``Three-tenths as full'' (\textit{mittsu j\=ubun no kiwameshi}) means
taking three parts of profit and counting that as ten parts---that is,
full. In other words, do not try to take ten-tenths profit; be satisfied
with three-tenths. If you try to take ten-tenths profit, you may instead
lose your gains. Be content with three-tenths profit, and yield the
remaining seven-tenths to others.

\bigskip

\textbf{(2)} Enter small, take profit thick. This is the cardinal principle of
the market. To enter big and take a small profit is highly dangerous
with little gain. To enter small and wait for profits to grow is safe
and yields great returns.

\bigskip

Furthermore, the timing of entry and the timing of exit
(\jt{手仕舞}{teshimai}) are equally important. If you err in timing
your entry, you incur losses; if you err in timing your exit, you
forfeit profits. Enter with caution; exit with decisiveness.

When entering, enter at the time when you still think it is too early.
When exiting, exit at the time when you still think it is too early.
If you enter too late, profits will be small; if you exit too late,
losses will be great.

\bigskip

\textit{Seed:} The method of entering in two parts is a way to test
whether one's forecast is correct or not. First enter with one part; if
there is profit, add the remainder. If there is loss, stop with the
smaller loss. He who follows this method will never suffer a great loss.

\bigskip

In taking profit, one must not act out of greed. The market is a public
thing under heaven. To try to exhaust all profit for oneself alone is to
go against the natural order. He who knows that three-tenths profit is
enough will sustain himself in the market for a long time.

The purpose of this chapter is to teach restraint in entering positions,
moderation in taking profit, and conservative trading. Engrave it well
upon your heart.

% =============================================================================
% UPPER VOLUME CHAPTER 15
% =============================================================================
\section*{Teaching 15: A Great Rise Being a Great Success}
\addcontentsline{toc}{section}{Teaching 15: A Great Rise Being a Great Success}

\begin{center}
\jp{第十五章　高上は大首尾の事}
\end{center}

\noindent Selling when prices have risen greatly and taking
a large profit is called a \textit{great success} (\jt{大首尾}{\=oshubi}).
A great success is what every trader desires, but it is not easily
obtained.

\bigskip

\textit{Seed:} To achieve a great success, one must first buy and stock
up at the bottom, then hold for a long time and wait for the ceiling. To
buy at the bottom, one must have the resolve to buy when everyone else
is selling. To sell at the ceiling, one must have the resolve to sell
when everyone else is buying. This is the secret of the great success.

However, identifying the ceiling is an extremely difficult task. One may
sell thinking it is the ceiling, only for prices to rise further. At
such times, be content with three-tenths profit as described before, and
do not force yourself to capture the very top.

Furthermore, a market that has risen greatly may fall suddenly. At such
times, exit quickly. A decline following a great rise is sharp and
severe. If one hesitates and fails to exit, one will not only lose
profits but may incur losses as well.

\bigskip

When one has achieved a great success (\jp{大首尾}, \textit{\=oshubi})
in a rising market, one should exit immediately. If one clings to
profits and fails to exit, one will instead incur losses. One should
consider a single great success sufficient. To try to achieve a great
success two or three times in succession is greed. The greedy will
certainly fail. Do not forget this principle.

% =============================================================================
% UPPER VOLUME CHAPTER 16
% =============================================================================
\chapter{Short Teachings on Market Patterns}
\markboth{Short Teachings on Market Patterns}{}

\section*{Teaching 16: Markets That Top Out, Bottom Out, and Turn to Falling Prices}
\addcontentsline{toc}{section}{Teaching 16: Markets That Top Out, Bottom Out, and Turn to Falling Prices}

\begin{center}
\jp{第十六章　天井底落値相場なる事}
\end{center}

\noindent To identify the ceiling (\jp{天井}, \textit{tenj\=o})
and bottom (\jp{底}, \textit{soko}) is the ultimate secret of the way of
the market. One sells thinking it is the ceiling, yet prices rise
further; one buys thinking it is the bottom, yet prices fall further.
This shows the extreme difficulty of identifying tops and bottoms.

\bigskip

\textbf{(1)} Three times the ceiling is tested without prices rising further;
three times the bottom is tested without prices falling further. When
this happens, the ceiling and bottom are confirmed. Wait for these three
tests before trading.

\bigskip

\textit{Seed:} A market that has hit its ceiling and turns to declining
prices falls slowly at first, then sharply later. He who sells at the
beginning of the decline earns great profit. He who sells at the
midpoint earns little. He who sells at the end, when the bottom is
already near, earns nothing. Therefore, in a falling market, it is
essential to sell early.

% =============================================================================
% UPPER VOLUME CHAPTER 17
% =============================================================================
\section*{Teaching 17: Five or Six Rallies During the July--August Decline}
\addcontentsline{toc}{section}{Teaching 17: Five or Six Rallies During the July--August Decline}

\begin{center}
\jp{第十七章　七八月天建五六ダケ引立の事}
\end{center}

\noindent July and August are the season when new-crop rice
comes to market, so prices normally fall. However, during this time
there are five or six rallies (\jt{引立}{hikitate}). A rally means a
temporary rise in prices. There is a method for profiting by riding
these rallies. However, since these rallies are temporary, one must not
chase them too far.

\bigskip

\textbf{(1)} During the declining market of July and August, know that there
will be five or six bounces, and at each bounce take a little profit. Do
not try to take large profits.

% =============================================================================
% UPPER VOLUME CHAPTER 18
% =============================================================================
\section*{Teaching 18: Extended Buying During Downward Momentum and Pullback}
\addcontentsline{toc}{section}{Teaching 18: Extended Buying During Downward Momentum and Pullback}

\begin{center}
\jp{第十八章　十二三月延買下勢引上の事}
\end{center}

\noindent From December through January, when buying on
margin (\jt{延買}{nobekai}) amid downward momentum, one should wait for
the opportunity to pull back upward. Even within a downtrend, when the
bottom has been tested three times without breaking, that is a good
opportunity for extended buying. If one buys at this time, prices often
pull upward by the New Year.

However, in years when the overall trend is downward, this method may
not work. Assess the overall trend before applying it.

% =============================================================================
% UPPER VOLUME CHAPTER 19
% =============================================================================
\section*{Teaching 19: Buying in July--August, Carrying Over, and Reaching the Top by December}
\addcontentsline{toc}{section}{Teaching 19: Buying in July--August, Carrying Over, and Reaching the Top by December}

\begin{center}
\jp{第十九章　七八月月越十二月を以て又天上りの事}
\end{center}

\noindent This is the method of buying rice in July--August,
carrying it over the months (\jt{月越}{tsukigoshi}) until December, and
waiting for the ceiling. This method---buying at summer lows and selling
at winter highs---can yield great profits. However, to carry a position
for six months requires ample capital.

Furthermore, if prices fall during the carry-over period, do not lose
heart---hold on.

% =============================================================================
% UPPER VOLUME CHAPTER 20
% =============================================================================
\section*{Teaching 20: Carrying a Bottom Position Over the Year-End into the New Year}
\addcontentsline{toc}{section}{Teaching 20: Carrying a Bottom Position Over the Year-End into the New Year}

\begin{center}
\jp{第二十章　底より越年正月の事}
\end{center}

\noindent This is the method of buying at the bottom,
carrying the position over the year-end, and taking profit in the New
Year. Rice that bottoms at year-end often rises in January. This method
involves buying during the cheap year-end period and selling during the
high New Year period.

Therefore, during November and December, identify the bottom and buy. If
prices rise in January, exit.

% =============================================================================
% UPPER VOLUME CHAPTER 21
% =============================================================================
\section*{Teaching 21: Extended High Prices in a Rising Market}
\addcontentsline{toc}{section}{Teaching 21: Extended High Prices in a Rising Market}

\begin{center}
\jp{第廿一章　上相場長値の事}
\end{center}

\noindent In a rising market, prices sometimes continue
upward for an extended period. During such an extended rally
(\jt{長値}{naganei}), one should not exit easily. An extended rally
reflects strong momentum---even if prices fall slightly, they will rise
again.

\bigskip

\textit{Seed:} To know when an extended rally is ending, observe the
change in crowd sentiment (\jp{人氣}, \textit{ninki}). As long as
everyone wants to buy, the rally continues. When the crowd grows weary
of buying, the rally ends.

% =============================================================================
% UPPER VOLUME CHAPTER 22
% =============================================================================
\section*{Teaching 22: The Tenjin Standstill from January to March--April}
\addcontentsline{toc}{section}{Teaching 22: The Tenjin Standstill from January to March--April}

\begin{center}
\jp{第廿二章　正月より三四月迄天神持の事}
\end{center}

\noindent From January through March or April, the market
sometimes ceases to move. This is called the \textit{Tenjin standstill}
(\jt{天神持合}{tenjin mochiai}). During a Tenjin standstill, there is
little profit in either selling or buying. At such times, keep your
hands out of the market and wait quietly.

\bigskip

\textbf{(1)} After a Tenjin standstill, the market will certainly make a big
move. Whether it moves up or down depends on the overall trend. If one
catches this big move, one can earn great profit.

If you desire great profit, endure patiently during the standstill and
ride the move when it begins. Piling up small trades during the
standstill earns little profit while fees mount.

\begin{hayasakanote}
The \textit{Tenjin standstill} refers to the market stagnation around
the time of the \=Osaka Tenjin festival.
\end{hayasakanote}

% =============================================================================
% UPPER VOLUME CHAPTER 23
% =============================================================================
\section*{Teaching 23: The May--June Settlement Month Decline}
\addcontentsline{toc}{section}{Teaching 23: The May--June Settlement Month Decline}

\begin{center}
\jp{第廿三章　五六月節月場達下の事}
\end{center}

\noindent May and June are settlement months
(\jt{節月}{setsugetsu}), a time when the market tends to fall.
Settlement months are the months for settling rice contracts; since
positions held on margin must be settled, there is heavy selling.

\bigskip

\textbf{(1)} In May and June, wheat also comes to market, increasing
substitutes for rice and reducing rice demand---so prices fall.

During this season, hold back on buying and look for opportunities to
sell. However, in famine years, prices may not fall even in May--June.
Assess the harvest quality of the year before acting.

% =============================================================================
% UPPER VOLUME CHAPTER 24
% =============================================================================
\section*{Teaching 24: Achieving Great Success in a Rising Market}
\addcontentsline{toc}{section}{Teaching 24: Achieving Great Success in a Rising Market}

\begin{center}
\jp{第廿四章　上相場の大首尾しの事}
\end{center}

\noindent To achieve a great success in a rising market, buy
at the bottom and hold through to the ceiling. Do not exit along the
way. To exit satisfied with a small profit is to let the great success
escape.

\bigskip

\textbf{(1)} However, since identifying the ceiling is extremely difficult, do
not forget the principle of being content with three-tenths profit. Do
not incur losses by trying to capture a great success.

\bigskip

\textit{Seed:} Great success in a rising market depends on two things:
patience (\jt{忍耐}{nintai}) and decisiveness (\jt{決斷}{ketsudan}).
Patience is the power to wait for profits to grow. Decisiveness is the
power to sell at the ceiling.

% =============================================================================
% UPPER VOLUME CHAPTER 25
% =============================================================================
\section*{Teaching 25: The Principles and Rules of Position Management}
\addcontentsline{toc}{section}{Teaching 25: The Principles and Rules of Position Management}

\begin{center}
\jp{第廿五章　仕合取込方理則の事}
\end{center}

\noindent In managing positions, there is both \textit{ri}
(\jp{理}, reason/principle) and \textit{soku} (\jp{則}, rule/method).
\textit{Ri} means judging by rational principle. \textit{Soku} means
acting according to established rules. He who possesses both reason and
rules wins in the market.

\bigskip

\textbf{(1)} Reason alone without knowing rules, or rules alone without knowing
reason---neither suffices. To know both together is the essential point
of the way of the market.

% =============================================================================
% UPPER VOLUME CHAPTER 26
% =============================================================================
\section*{Teaching 26: The Three Market Modes}
\addcontentsline{toc}{section}{Teaching 26: The Three Market Modes}

\begin{center}
\jp{第廿六章　三つ相場離□しの事}
\end{center}

\noindent The market ``separating into three'' means the
three forms: rising, falling, and consolidation (\jt{持合ひ}{mochiai}).
To distinguish among these three forms is a matter of great importance
in the way of the market. When rising, buy; when falling, sell; when
consolidating, rest. He who observes these three rules wins in the
market.

\bigskip

\textit{Seed:} This principle is simple in statement yet difficult in
practice. The challenge lies not in knowing the three modes but in
correctly identifying which mode the market is currently in. Study the
overall trend, observe crowd sentiment, and only then determine the
mode. He who acts on a wrong determination of mode will certainly incur
losses.

% =============================================================================
% UPPER VOLUME CHAPTER 27
% =============================================================================
\section*{Teaching 27: The True Third Wave and the Fourth--Fifth--Sixth--Seventh Waves}
\addcontentsline{toc}{section}{Teaching 27: The True Third Wave and the Fourth--Fifth--Sixth--Seventh Waves}

\begin{center}
\jp{第廿七章　正三番違合四五六七の事}
\end{center}

\noindent The ``true third'' (\jt{正三番}{sh\=o sanban})
refers to the momentum shift at the third wave. The ``four-five-six-seven''
refers to the fourth, fifth, sixth, and seventh waves. The market moves
up and down like waves, and to know the number of those waves is
essential to the way of the market.

\bigskip

\textbf{(1)} At the third wave, there is sometimes a major move. The fourth and
fifth waves bring smaller waves in succession. By the sixth and seventh
waves, the momentum begins to wane. Know the movement of these waves
and trade accordingly.

In the March market, watch the movement of the true third wave and then
act. From April through July, ride the smaller waves and take profit
little by little.

% =============================================================================
% UPPER VOLUME CHAPTER 28
% =============================================================================
\chapter{The Trader's Discipline}
\markboth{The Trader's Discipline}{}

\section*{Teaching 28: November Bottoms and Ceilings}
\addcontentsline{toc}{section}{Teaching 28: November Bottoms and Ceilings}

\begin{center}
\jp{第廿八章　霜月底天井の事}
\end{center}

\noindent The frost month (\jt{霜月}{shimotsuki}---November)
is the month when the market reaches either its bottom or its ceiling.
This is the time when the year's harvest is determined and the outlook
for the following year takes shape, and so the market may move greatly.

\bigskip

\textbf{(1)} In an abundant year, November marks the bottom. In a famine year,
November marks the ceiling. Know this principle and pay attention to the
November market.

A market that has bottomed in November will rise heading into the New
Year. A market that has topped in November will fall heading into spring.
This pattern is connected to the principle of the three pivotal months
(\jp{三位の月}, \textit{sanmi no tsuki}) discussed in the preceding
chapter; consider them together.

% =============================================================================
% UPPER VOLUME CHAPTER 29
% =============================================================================
\section*{Teaching 29: Not Wagering Beyond One's Means}
\addcontentsline{toc}{section}{Teaching 29: Not Wagering Beyond One's Means}

\begin{center}
\jp{第廿九章　大金不相應勝負の事}
\end{center}

\noindent In the market, one must not wage contests
disproportionate to one's capital. He who commits large sums seeking a
big win will certainly suffer a big loss. Wage contests suited to your
strength.

\bigskip

\textbf{(1)} A contest disproportionate to one's means is like a man with the
strength of ten waging a contest of a hundred. Such a man, upon a single
failure, cannot recover. If you have the strength of ten, wage the
contest of ten. This is the iron rule of the way of the market.

Do not borrow money to trade. Do not gamble your family estate on the
market. Trade only with your surplus capital. Do not lose large sums and
destroy yourself.

% =============================================================================
% UPPER VOLUME CHAPTER 30
% =============================================================================
\section*{Teaching 30: The Outcome of Accumulating Buys During a Consolidation}
\addcontentsline{toc}{section}{Teaching 30: The Outcome of Accumulating Buys During a Consolidation}

\begin{center}
\jp{第三十章　持合合買首尾の事}
\end{center}

\noindent The method of gradually buying during a
consolidation (\jp{持合}, \textit{mochiai}) is called \jt{合買}{g\=ogai}
(combined buying). During the consolidation, buy at the low points and
sell a little at the high points. This method accumulates small profits
over time.

\bigskip

\textbf{(1)} If the market breaks upward after the consolidation, one earns
great profits. But if it breaks downward, one may incur losses.
Therefore, during a consolidation, buy only gradually---do not commit
your full strength.

\bigskip

\textit{Seed:} Consolidation is the market's resting time. You too
should rest and prepare for the next move. Piling up forced trades
during a consolidation earns no profit while fees mount. Waiting quietly
is essential.

% =============================================================================
% UPPER VOLUME CHAPTER 31
% =============================================================================
\section*{Teaching 31: Pyramiding Buys to Turn a Win into a Big Win}
\addcontentsline{toc}{section}{Teaching 31: Pyramiding Buys to Turn a Win into a Big Win}

\begin{center}
\jp{第三十一章　可買入合段勝大勝の事}
\end{center}

\noindent This is the method of combining entry levels
(\jt{合段}{g\=odan}) to turn a win into a big win. Buy a little at the
bottom; when prices rise, buy a little more. This is called pyramiding
(\jp{合段}, \textit{g\=odan}). By adding to a winning position, profits
grow greatly.

\bigskip

\textbf{(1)} However, one must confirm that the overall trend is upward before
adding. To add when the trend is uncertain is dangerous.

\bigskip

\textit{Seed:} The pyramiding method should be used only after
confirming the bottom. Use this method only when the bottom is firm. To
pyramid in the middle of a falling market only magnifies losses.

% =============================================================================
% UPPER VOLUME CHAPTER 32
% =============================================================================
\section*{Teaching 32: The Three-Month Influence and Pullback Outcomes}
\addcontentsline{toc}{section}{Teaching 32: The Three-Month Influence and Pullback Outcomes}

\begin{center}
\jp{第三十二章　三ヶ月影響と引上首尾の事}
\end{center}

\noindent The ``three-month influence'' means that the
market's movement three months ago affects today's market. A market
that bottomed three months ago is likely to rise now. A market that
topped three months ago is likely to fall now. Knowing this three-month
cycle is a matter of great importance in the way of the market.

\bigskip

\textbf{(1)} ``Pullback outcomes'' refers to the results of a market that has
pulled upward over three months. A market that has risen greatly over
three months will experience a correction afterward.

% =============================================================================
% UPPER VOLUME CHAPTER 33
% =============================================================================
\section*{Teaching 33: Resting for Forty-Seven Days After a Stalled Position}
\addcontentsline{toc}{section}{Teaching 33: Resting for Forty-Seven Days After a Stalled Position}

\begin{center}
\jp{第三十三章　仕掛離れ四十七日休の事}
\end{center}

\noindent When a position one has entered fails to move as
expected, set a limit of forty-seven days and rest. If there is no
profit within forty-seven days, exit the position and rest.

\bigskip

\textbf{(1)} Resting is also important in the market. Those who trade
constantly grow weary in body and mind, and often misjudge. Take a
forty-seven-day rest, refresh your mind, and only then enter a new
position.

% =============================================================================
% UPPER VOLUME CHAPTER 34
% =============================================================================
\section*{Teaching 34: Regional Procurement with Consideration for Costs and Connections}
\addcontentsline{toc}{section}{Teaching 34: Regional Procurement with Consideration for Costs and Connections}

\begin{center}
\jp{第三十四章　國仕相買入判分料緣心國の事}
\end{center}

\noindent In regional procurement, one should separate the
assessment, examine the cost amounts, and apply one's judgment. When
buying rice from distant provinces, calculate the transport costs and
investigate the harvest quality of that province, and only then buy.

\bigskip

\textbf{(1)} Procurement from distant provinces requires many days, and one must
be prepared for market changes during that time. Rice from nearby
provinces requires fewer days and is easier to profit from.

% =============================================================================
% UPPER VOLUME CHAPTER 35
% =============================================================================
\section*{Teaching 35: Following the Predecessors Even When the World Declines}
\addcontentsline{toc}{section}{Teaching 35: Following the Predecessors Even When the World Declines}

\begin{center}
\jp{第三十五章　天下沈むとて先達の事}
\end{center}

\noindent Even when the world declines, one should follow
the teachings of the predecessors (\jt{先達}{sendatsu}). The way of the
market does not change across the ages. Even amid upheaval in the world,
the principles of the market remain the same. He who follows the
predecessors' teachings and acts in accordance with reason can preserve
himself in any circumstances.

\bigskip

\textbf{(1)} In times of turmoil, everyone is frightened and sells. He who at
such times acts without fear, following reason, will earn great profit.
The predecessors teach: ``He who walks the road that others do not walk
is the one who profits.'' Follow this teaching.

Furthermore, read the writings of the predecessors and learn what they
did. Old Man Honma S\=oky\=u is the great predecessor of this way, and
his teachings are worth a thousand pieces of gold.

\bigskip

\textit{Seed:} The predecessors' teachings do not grow old even as times
change. Reason (\jt{道理}{d\=ori}) is eternal and unchanging. To
understand this reason and apply it to one's own trading---this is
essential.

He who profits in times of turmoil is he who follows the predecessors'
teachings and acts in accordance with reason---and no other. He who is
confused by the uproar of the world and lets his heart grow disordered
will certainly incur losses.

% =============================================================================
% UPPER VOLUME CHAPTER 36
% =============================================================================
\chapter{Reading the Market}
\markboth{Reading the Market}{}

\section*{Teaching 36: Dips Within a Rising Market}
\addcontentsline{toc}{section}{Teaching 36: Dips Within a Rising Market}

\begin{center}
\jp{第三十六章　上の内下相場の事}
\end{center}

\noindent Even within a rising market, there are temporary
declines. These are called ``dips within the rise'' (\textit{ue no uchi
no sage}). A dip within the rise is merely a temporary correction within
an upward trend---one must not sell on such a dip. Rather, a dip is
precisely the time to buy.

To distinguish between a dip within the rise and a genuine decline is
essential to the way of the market. A dip within the rise is small in
magnitude and short in duration. A genuine decline is large in magnitude
and long in duration.

\bigskip

\textbf{(2)} Furthermore, when everyone is frightened and sells during a dip
within the rise, the overall trend remains upward, so prices will
immediately rise again. He who buys at such a time will earn profit. He
who fears a dip within the rise loses sight of the overall trend.

\bigskip

\textbf{(3)} Dips within the rise normally occur up to three times. The first is
shallow, the second is deep, and the third is shallow again. After the
third dip, there is often a major advance.

% =============================================================================
% UPPER VOLUME CHAPTER 37
% =============================================================================
\section*{Teaching 37: The Right Mental Posture During Unfavorable Periods}
\addcontentsline{toc}{section}{Teaching 37: The Right Mental Posture During Unfavorable Periods}

\begin{center}
\jp{第三十七章　不利節の勢心勢方の事}
\end{center}

\noindent How to hold one's mind during unfavorable periods
is a matter of great importance in the way of the market. In unfavorable
times, first calm your mind and do not become frantic. He who
frantically tries to recoup will invite still deeper losses.

\bigskip

\textbf{(1)} In unfavorable times, reassess the overall trend and reflect on
the errors in your own judgment. Correct the posture of your mind, and
only then enter anew. To enter a position with a disordered mind will
invite further losses.

\bigskip

\textit{Seed:} In unfavorable times, rest awhile and restore your
composure. Resting is also part of the way of the market. During a
losing streak, you must rest at least once and refresh your mind.

% =============================================================================
% UPPER VOLUME CHAPTER 38
% =============================================================================
\section*{Teaching 38: The Right Mental Posture for Each Season}
\addcontentsline{toc}{section}{Teaching 38: The Right Mental Posture for Each Season}

\begin{center}
\jp{第三十八章　時道の勢心勢方の事}
\end{center}

\noindent The ``way of the seasons'' (\textit{toki no michi})
refers to the market's movements in accordance with the time of year.
Spring is the time for rising; summer is the time for falling; autumn is
the time for bottoms; winter is the time for ceilings. Know this
seasonal way and align your mental posture accordingly.

\bigskip

\textit{Seed:} He who follows the way of the seasons gains profit; he
who goes against the way of the seasons incurs losses. Buy in spring,
sell in summer, rest in autumn, prepare in winter. This is the way of
the four seasons. He who follows this way will not commit a great error
in the market.

However, in some years this rule does not hold. Observe the overall
trend and adapt flexibly.

% =============================================================================
% UPPER VOLUME CHAPTER 39
% =============================================================================
\section*{Teaching 39: Studying Writings on the Economy}
\addcontentsline{toc}{section}{Teaching 39: Studying Writings on the Economy}

\begin{center}
\jp{第三十九章　經濟書かむら事}
\end{center}

\noindent To learn the way of the market, one should read
books on the economy (\jt{經濟}{keizai}). Since the price of rice moves
according to the economy of the realm, one must understand the
principles of the economy.

\bigskip

\textbf{(1)} Learn the harvest conditions in the provinces, the state of the
economy, the circulation of gold and silver---and use this knowledge to
judge the overall trend of the market. To trade relying only on the
prices before one's eyes is like a blind man walking a road in the dark
of night.

\bigskip

\textit{Seed:} Read the writings of the predecessors, learn what they
did, and apply it to your own trading---this is essential.

% =============================================================================
% UPPER VOLUME CHAPTER 40
% =============================================================================
\section*{Teaching 40: Broken Bottoms and Broken Ceilings}
\addcontentsline{toc}{section}{Teaching 40: Broken Bottoms and Broken Ceilings}

\begin{center}
\jp{第四十章　底ぬけ乃天井ぬけのの事}
\end{center}

\noindent A``broken bottom'' (\textit{sokonuke}) means that
prices break below what was thought to be the bottom and fall further
still. A ``broken ceiling'' (\textit{tenj\=onuke}) means that prices
break above what was thought to be the ceiling and rise further still. A
broken bottom or ceiling marks the beginning of a major market upheaval.

When a bottom breaks, everyone is frightened and sells. Yet after a
broken bottom, the \textit{major bottom} (\jt{大底}{\=ozoko}) may be
reached. When a ceiling breaks, everyone rejoices and buys. Yet after a
broken ceiling, the \textit{major ceiling} may be reached.

\bigskip

\textit{Seed:} When bottoms or ceilings break, do not let your heart be
moved---observe the overall trend calmly. The major bottom after a
broken bottom is an excellent opportunity for buying. The major ceiling
after a broken ceiling is an excellent opportunity for selling.

Furthermore, broken bottoms and ceilings occur during major upheavals
that come once every three to five years. He who does not let these
upheavals escape and takes his profit will obtain great wealth.

\bigskip

The signs of a broken bottom are as follows:

\begin{enumerate}
\item Prices break below the previous bottom and continue to fall.
\item Everyone is frightened and rushes to sell.
\item Trading volume (\jt{出來高}{dekidaka}) increases greatly.
\end{enumerate}

When all three of these conditions are present, it is a broken bottom.
However, after a broken bottom, prices may hit a major bottom and
reverse. That reversal is an excellent opportunity for buying.

The signs of a broken ceiling are as follows:

\begin{enumerate}
\item Prices break above the previous ceiling and continue to rise.
\item Everyone rejoices and rushes to buy.
\item Trading volume increases greatly.
\end{enumerate}

When all three conditions are present, it is a broken ceiling.

\bigskip

The major upheaval following a broken bottom or ceiling may last several
months. During this upheaval, do not close your position after a single
trade. Hold as long as the momentum continues.

However, when momentum fades, exit immediately. The fading of momentum
is known by the size of price movements becoming smaller. When a market
that had been moving greatly begins to move only slightly, it is proof
that momentum has faded.

This chapter sets forth the principles for navigating major market
upheavals---engrave it well upon your heart.

% =============================================================================
% UPPER VOLUME CHAPTER 41
% =============================================================================
\section*{Teaching 41: Profiting from an Extended Rise}
\addcontentsline{toc}{section}{Teaching 41: Profiting from an Extended Rise}

\begin{center}
\jp{第四十一章　上首尾長値の事}
\end{center}

\noindent Asuccessful outcome (\jt{上首尾}{j\=oshubi}) means
having gained profit from buying. \textit{Naganei}
(\jt{長値}{naganei}) means prices continuing to rise over a long period.
When one has a successful position in an extended rally, one should not
exit easily.

\bigskip

\textbf{(1)} In an extended rally, even if prices fall slightly, they will rise
again. Wait for others to sell, then add to your position. The end of an
extended rally is known by changes in crowd sentiment (\jp{人氣},
\textit{ninki}).

\bigskip

\textit{Seed:} The signs that an extended rally is ending are as follows:

\begin{enumerate}
\item The magnitude of each advance grows smaller.
\item The crowd becomes weary of buying.
\item Trading volume declines.
\end{enumerate}

When these three signs are visible, the end of the extended rally is
near. That is the time to exit.

\bigskip

\textit{Seed:} If one fails to notice the end of an extended rally and
continues holding, profits will vanish in an instant and turn to losses.
Identify the signs that the rally is ending, and do not miss the time to
exit.

% =============================================================================
% UPPER VOLUME CHAPTER 42
% =============================================================================
\section*{Teaching 42: It Takes Eighteen Years to Become a Full-Fledged Trader}
\addcontentsline{toc}{section}{Teaching 42: It Takes Eighteen Years to Become a Full-Fledged Trader}

\begin{center}
\jp{第四十二章　相場是年十八が一人前の事}
\end{center}

\noindent To master the way of the market requires years
and months. Only after eighteen years of training does one become a
full-fledged (\jt{一人前}{ichininmae}) market trader. In the first three
years, one incurs losses and learns how fearsome the market is. In the
next five years, one earns small profits and learns the methods. In the
following ten years, one cultivates the eye for seeing the overall
trend.

\bigskip

\textbf{(2)} The way of the market is deep and far-reaching. It cannot be
mastered overnight. He who impatiently seeks great profit will certainly
fail. Cultivate it over years and months, study day and night, and only
then will you attain its subtlety.

He who has not undergone these eighteen years of training should not
enter the way of the market. Begin from youth and accumulate training
over the years. The market is a lifelong path. Do not be dazzled by
momentary profits; take the long view.

\bigskip

\textit{Seed:} The way of the market resembles the way of the martial
arts (\jt{武藝}{bugei}). In the martial arts too, one receives
certification (\jt{免許}{menkyo}) only after years of training. The
market is the same. Consider eighteen years of training as your
certification.

% =============================================================================
% UPPER VOLUME CHAPTER 43
% =============================================================================
\chapter{Practical Methods}
\markboth{Practical Methods}{}

\section*{Teaching 43: The Right Month to Buy, Price Entry, and Mental Posture}
\addcontentsline{toc}{section}{Teaching 43: The Right Month to Buy, Price Entry, and Mental Posture}

\begin{center}
\jp{第四十三章　買月値段入心勢方の事}
\end{center}

\noindent The ``buying month'' (\textit{kaigetsu}) refers to
the month when one should buy. Know the proper way to enter at a given
price and the proper mental posture (\jt{心勢}{shinsei}), and only then
buy. The months for buying are September and October, when prices are at
the bottom. At this time, steady your mind and enter at a low price.

``Mental posture'' (\jp{心勢}, \textit{shinsei}) refers to how one holds
one's mind when approaching the market. Be calm and unhurried, acting in
accordance with reason. When your mind is disordered, do not trade.

\bigskip

\textit{Seed:} Know the buying month, identify the right price entry,
and align your mental posture---only when all three are in place should
you buy. If any of the three is missing, do not buy.

% =============================================================================
% UPPER VOLUME CHAPTER 44
% =============================================================================
\section*{Teaching 44: The Mental Error of Buying in a Rising Market}
\addcontentsline{toc}{section}{Teaching 44: The Mental Error of Buying in a Rising Market}

\begin{center}
\jp{第四十四章　上相場にて可買心違方の事}
\end{center}

\noindent When one tries to buy in a rising market, there
is a tendency toward mental error (\jt{心違}{kokoroechigai}). To see a
market that has already risen and think it will rise further, and buy on
that basis---that is a mental error.

A market that has already risen has, because it has already risen, little
room to rise further. To buy at such a time is the same as buying the
ceiling (\jp{天井}).

\bigskip

\textit{Seed:} The right time to buy in a rising market is during a dip
within the rise (\textit{ue no uchi no sage}). Buying when prices have
temporarily dipped is an easy way to profit. Buying when prices have
risen is both unprofitable and dangerous. Do not forget this principle.

% =============================================================================
% UPPER VOLUME CHAPTER 45
% =============================================================================
\section*{Teaching 45: The Misjudged Outcome and Wrong Momentum in a Falling Market}
\addcontentsline{toc}{section}{Teaching 45: The Misjudged Outcome and Wrong Momentum in a Falling Market}

\begin{center}
\jp{第四十五章　下相場にて已首達違勢方の事}
\end{center}

\noindent In a falling market, to see that prices have
already fallen and think the bottom has been reached and buy---this is
sometimes premature. The momentum of a falling market can break below
what one thought was the bottom and fall further still.

It is essential to distinguish between where an outcome reaches its
conclusion and where the momentum is heading. Before concluding that a
bottom has been reached, test it three times and only then make your
determination.

\bigskip

\textit{Seed:} In a falling market, do not rush to buy at the bottom.
Confirm that the bottom is firm, and only then buy. This, as stated in
earlier chapters, is a matter of great importance in the way of the
market.

% =============================================================================
% UPPER VOLUME CHAPTER 46
% =============================================================================
\section*{Teaching 46: Comparing Methods and Their Effectiveness}
\addcontentsline{toc}{section}{Teaching 46: Comparing Methods and Their Effectiveness}

\begin{center}
\jp{第四十六章　藥は達方比可りの事}
\end{center}

\noindent In the way of the market, ``medicine''
(\textit{kusuri}) refers to the methods for gaining profit. ``Reach''
(\textit{tachigata}) refers to where those methods arrive. Compare them
and choose the one that yields the greater profit.

\bigskip

\textit{Seed:} There are various methods for taking profit. Compare the
strengths and weaknesses of each method, and use the one that suits you.
Do not rely on a single method alone; change your method to suit the
time and circumstances.

The way of the market is not one single thing. Know various methods and
use them as the occasion demands. To be rigid about a single method is
not the way of the market. Be like water, changing form to suit the
vessel. This is the subtlety of the market.

\bigskip

\textit{Seed:} Just as medicine becomes poison when taken in excess, a
method for taking profit becomes a source of loss when overused. To
know the right measure is essential.

% =============================================================================
% UPPER VOLUME CHAPTER 47
% =============================================================================
\section*{Teaching 47: Big Money Entering the Contest Between Sellers and Buyers}
\addcontentsline{toc}{section}{Teaching 47: Big Money Entering the Contest Between Sellers and Buyers}

\begin{center}
\jp{第四十七章　賣方買引大金値}
\end{center}

\noindent When big money enters the contest between sellers
and buyers, the market moves greatly. The market moves in the direction
where big money enters. When there is heavy big-money selling, prices
fall; when there is heavy big-money buying, prices rise.

\bigskip

\textbf{(1)} To detect the movement of big money is a matter of great
importance in the way of the market. To chase after big money that has
already entered is sometimes too late. Detect the signs that big money
is about to enter, and move ahead of it.

\bigskip

\textit{Seed:} The movement of big money is known by the increase or
decrease in trading volume (\jp{出來高}, \textit{dekidaka}). When volume
increases greatly, it is proof that big money has entered.

% =============================================================================
% UPPER VOLUME CHAPTER 48
% =============================================================================
\section*{Teaching 48: Waiting Years Beyond the Three Markets}
\addcontentsline{toc}{section}{Teaching 48: Waiting Years Beyond the Three Markets}

\begin{center}
\jp{第四十八章　三田の外の數年守の事}
\end{center}

\noindent ``Waiting years beyond the three markets'' means
waiting for several years outside the three trading seasons. The great
opportunities of the market come only once in several years. Wait for
such an opportunity, and then move decisively.

The first principle is waiting. The market is a path where the patient
win. He who moves without waiting gains small profits but forfeits great
ones.

Within the span of several years, there are great upheavals in harvest
quality. In famine years, rice prices rise greatly; in abundant years,
they fall greatly. These great upheavals are the prime opportunity to
take great profit.

\bigskip

Once every three to five years there is a great opportunity. In
preparation for that opportunity, spend your ordinary time practicing
with small trades, accumulating capital, and building your strength.

When the great opportunity comes, move with all your strength. But ``all
your strength'' must not exceed the range of your own capital.

\bigskip

\textit{Seed:} Like a field---one plows for several years and then reaps
the harvest. The market is the same. After years of training and
patience, one reaps great profits.

\bigskip

\textit{Seed:} Furthermore, the market is not an amusement
(\jt{道樂}{d\=oraku}). It is a livelihood (\jt{生業}{seigy\=o}). He who
trades as a livelihood should always keep his heart right and act in
accordance with reason. He who trades as an amusement will certainly
destroy himself.

The market is a public thing under heaven. Do not covet profit for
yourself alone. He who acts in accordance with the natural order will
profit for a long time. He who goes against the natural order may gain
momentary profit, but will ultimately incur losses.

\bigskip

The way of the market is: cultivate the mind, make the body upright,
investigate principle, and only then gain profit. If the heart is not
right, one cannot gain profit. If the body is not upright, one cannot
investigate the way.

This chapter sets forth the foundation of the way of the market---engrave
it well upon your heart.

% =============================================================================
% UPPER VOLUME CHAPTER 49
% =============================================================================
\section*{Teaching 49: The Unsettled Momentum of the Summer Market}
\addcontentsline{toc}{section}{Teaching 49: The Unsettled Momentum of the Summer Market}

\begin{center}
\jp{第四十九章　夏中の諸時は違ぬらぬ勢}
\end{center}

\noindent Summer is a time when the market's momentum is
unsettled. Summer is the time of waiting for new-crop rice, and the
market tends to move erratically in either direction. At such times, do
not enter large positions.

In the summer market, amid the unsettled momentum, there are small
waves. Ride these small waves and take small profits. To enter large
positions seeking large profits in the summer market is dangerous.

\bigskip

\textit{Seed:} Summer is the market's resting time. Rest as well, and
prepare for the great upheaval of autumn.

% =============================================================================
% UPPER VOLUME CHAPTER 50
% =============================================================================
\section*{Teaching 50: The Mental Error of Entering Positions After a Win}
\addcontentsline{toc}{section}{Teaching 50: The Mental Error of Entering Positions After a Win}

\begin{center}
\jp{第五十章　勝方仕掛心違の事}
\end{center}

\noindent After a win, one is prone to mental error when
entering the next position. To enter aggressively on the heels of a
victory is the foundation of defeat. It is precisely after a win that
one should exercise restraint and hold back.

\bigskip

\textit{Seed:} After a win, the heart grows inflated and one is likely
to misjudge. It is precisely after a win that one should tighten the
reins of the heart and act with caution. After a winning streak, rest
once and refresh your mind.

He who is arrogant in victory will certainly lose. He who is humble will
win for a long time. He who knows this principle may be said to have
grasped the way of the market.

% =============================================================================
% UPPER VOLUME CHAPTER 51
% =============================================================================
\section*{Teaching 51: Buying at Seventy-Three Monme}
\addcontentsline{toc}{section}{Teaching 51: Buying at Seventy-Three Monme}

\begin{center}
\jp{第五十一章　七十三匁買の事}
\end{center}

\noindent Seventy-three \textit{monme} (\jp{匁}) refers to
a rice price of seventy-three monme per \textit{koku} (\jp{石}). This
price is a benchmark for cheap prices---when prices reach this level, it
is a good opportunity to buy.

Since seventy-three monme is a benchmark for bottom prices, when prices
approach this level, one should prepare to buy. However, in years of
severe famine, prices may break below seventy-three monme and fall
further still. At such times, confirm that the bottom is firm before
buying.

Furthermore, seventy-three monme is a price that comes only once every
three to five years. If one buys at this cheap price, one may earn great
profits. Wait for this cheap price over the course of several years, and
buy when it arrives. This is the method of ``waiting years beyond the
three markets'' described in the previous chapter.

\bigskip

\textit{Seed:} The seventy-three monme benchmark applies to the Sakata
rice market; in other locations the benchmark will differ. However, the
importance of establishing a price benchmark and waiting patiently for
it is the same everywhere.

\bigskip

He who would master the way of the market must first understand the
principle of why rice prices rise and fall. Seventy-three monme is a
benchmark for the bottom, yet prices can fall further still to thirty or
fifty monme. One hundred monme is a benchmark for the ceiling, yet
prices can rise further still to one hundred twenty or one hundred fifty
monme.

Thus, while benchmarks exist, one must not trade solely on benchmarks.
Observe the overall trend, know the season, read the crowd's sentiment,
and only then trade.

Furthermore, if one buys at the seventy-three monme benchmark yet prices
fall to fifty monme, one should acknowledge the error in one's forecast
and exit quickly. Do not cling to a benchmark and let losses grow.

The market moves according to the time of heaven and the hearts of men;
it is difficult to measure by a fixed benchmark alone. Use benchmarks as
reference, but judge by the overall trend.

Rice prices rise and fall according to various circumstances, and cannot
be fixed in advance. Yet having a benchmark is important---to trade with
no benchmark at all is like a blind man walking a road.

% =============================================================================
% UPPER VOLUME CHAPTER 52
% =============================================================================
\chapter{Advanced Principles}
\markboth{Advanced Principles}{}

\section*{Teaching 52: The Winter-to-Spring Market}
\addcontentsline{toc}{section}{Teaching 52: The Winter-to-Spring Market}

\begin{center}
\jp{第五十二章　冬より春を通じ相場米相の事}
\end{center}

\noindent From winter through spring, rice prices generally
rise. Winter demand is high and inventories low---hence prices rise.
They continue rising through spring as long as demand persists. The
normal course is to buy from winter through spring and sell at the
May--June ceiling. However, in some years the ceiling comes in winter
and prices fall in spring. Know these exceptions; they depend on
harvest quality.

\bigskip

\textit{Seed:} The winter-spring market depends greatly on the harvest.
In abundant years, the rise is small. In famine years, the rise is
great. Assess the year's harvest before committing.

% =============================================================================
% UPPER VOLUME CHAPTER 53
% =============================================================================
\section*{Teaching 53: Unwinding in Three Parts After a Market Top}
\addcontentsline{toc}{section}{Teaching 53: Unwinding in Three Parts After a Market Top}

\begin{center}
\jp{第五十三章　天井投落直相三つ解きの事}
\end{center}

\noindent When the market has topped and begins to fall, one
should unwind (\jp{解く}, \textit{toku}) one's position in three parts.
Sell one-third when you believe the ceiling has been reached. Sell
another third when prices begin to fall. Sell the final third when
prices have fallen further. By dividing the exit into three parts, even
if you misjudge the ceiling, you avoid catastrophic loss.


\bigskip

\textit{Seed:} This three-part method applies to buying as well. Buy
one-third at the presumed bottom, another third as prices begin to rise,
and the remainder as the rise confirms.

% =============================================================================
% UPPER VOLUME CHAPTER 54
% =============================================================================
\section*{Teaching 54: Evaluating the High Price at a Ceiling}
\addcontentsline{toc}{section}{Teaching 54: Evaluating the High Price at a Ceiling}

\begin{center}
\jp{第五十四章　天井の高値上評の事}
\end{center}

\noindent To evaluate a ceiling's high price, compare it to
the previous ceiling. If higher than the previous ceiling, momentum is
strong. If lower, momentum has weakened. The ceiling's height depends on
the year's harvest quality and the strength of crowd sentiment.

Furthermore, a ceiling reached after a long steady rise tends to be
high. A ceiling reached after a sharp short rise, though high, tends to
fall back quickly.

% =============================================================================
% UPPER VOLUME CHAPTER 55
% =============================================================================
\section*{Teaching 55: Carrying Positions Past the June Settlement}
\addcontentsline{toc}{section}{Teaching 55: Carrying Positions Past the June Settlement}

\begin{center}
\jp{第五十五章　六月越し達の事}
\end{center}

\noindent Carrying positions past June (\textit{rokugatsu
goshi}) means holding rice through the June settlement month. June is a
settlement month with heavy selling, so holding through requires ample
capital. After June, prices may continue falling in July and August.
Without the strength to endure this decline, one should not attempt the
June carry-over. Use this method only in years when the overall trend
is upward.


% =============================================================================
% UPPER VOLUME CHAPTER 56
% =============================================================================
\section*{Teaching 56: Large and Small Rises Within Rising Markets}
\addcontentsline{toc}{section}{Teaching 56: Large and Small Rises Within Rising Markets}

\begin{center}
\jp{第五十六章　上は相場とても相場とて大小の事}
\end{center}

\noindent Within rising markets there are both large and
small rises. A large rise is driven by the overall trend and can last
months or more than a year. A small rise is driven by temporary
sentiment and lasts only days or weeks. For large rises, commit boldly.
For small rises, take modest profits and exit. To treat both the same is
a fundamental error. The same principle applies to falling
markets---there are large and small declines. Distinguishing between
them is essential.


% =============================================================================
% UPPER VOLUME CHAPTER 57
% =============================================================================
\section*{Teaching 57: Misjudging the Relationship Between Money and Rice}
\addcontentsline{toc}{section}{Teaching 57: Misjudging the Relationship Between Money and Rice}

\begin{center}
\jp{第五十七章　錢米判相心違の事}
\end{center}

\noindent Misjudging the relationship between money
(\jp{錢}, \textit{zeni}) and rice is common in trading. One judges the
money market by the rice market and vice versa. Understanding the
relationship between these two markets is essential. They do not always
move in concert---sometimes they move in opposite directions. Know this
discrepancy and correct your judgment accordingly.


% =============================================================================
% UPPER VOLUME CHAPTER 58
% =============================================================================
\section*{Teaching 58: Buying at the Water's Low Point}
\addcontentsline{toc}{section}{Teaching 58: Buying at the Water's Low Point}

\begin{center}
\jp{第五十八章　買の水下可買}
\end{center}

\noindent The ``water's low point'' for buying means the low price at
which one should buy. As water flows to the lowest point, so should one
buy at the lowest prices. To buy at high prices is to go against the
natural order, like water flowing uphill.


% =============================================================================
% UPPER VOLUME CHAPTER 59
% =============================================================================
\section*{Teaching 59: There Being Two Types of Extremes}
\addcontentsline{toc}{section}{Teaching 59: There Being Two Types of Extremes}

\begin{center}
\jp{第五十九章　極相に二つある事}
\end{center}

\noindent There are two types of price extremes
(\jt{極相}{gokus\=o}). One is the ceiling extreme, where upward
momentum exhausts itself and decline begins. The other is the bottom
extreme, where downward momentum exhausts itself and advance begins.
Identifying these two extremes is the ultimate secret of the way of the
market.


% =============================================================================
% UPPER VOLUME CHAPTER 60
% =============================================================================
\section*{Teaching 60: Paying Attention to Each Day's Extremes}
\addcontentsline{toc}{section}{Teaching 60: Paying Attention to Each Day's Extremes}

\begin{center}
\jp{第六十章　毎日の極は氣を付け事}
\end{center}

\noindent Pay attention to each day's extremes. Watch each day's high
and low and record them. When daily highs gradually climb higher, it is
proof of a rising market. When daily lows gradually sink lower, it is
proof of a falling market.

\bigskip
\textit{Seed:} Watching each day's extremes is the foundation of the
way of the market. The overall trend is known through the accumulation
of each day's small movements. Apply your mind to each day's movement,
and thereby judge the overall trend.

Keep a trading record (\jt{帳面}{ch\=omen}) and note each day's highs
and lows. If you look at the record, the overall trend becomes clear.
To trade without a record is like walking a road blindfolded.

% =============================================================================
% UPPER VOLUME CHAPTER 61
% =============================================================================
\section*{Teaching 61: Three Consecutive Wins with Big Money}
\addcontentsline{toc}{section}{Teaching 61: Three Consecutive Wins with Big Money}

\begin{center}
\jp{第六十一章　大入金三つ勝勢の事}
\end{center}

\noindent When one commits large capital and wins three times, the
overall trend is strong in one's favor. If you commit large capital
three times and profit all three times, the trend is on your side. In
such a case, continue holding.

However, after three wins, hold back a little. To bet heavily in the
flush of victory is the foundation of defeat. After three wins, rest
once and refresh your mind.

\bigskip
\textit{Seed:} Three consecutive wins are not coincidence. It is
because the overall trend is in your favor. Yet the overall trend
always changes, so after three wins one must not let down one's guard.

% =============================================================================
% UPPER VOLUME CHAPTER 62
% =============================================================================
\section*{Teaching 62: Holding from the Bottom Through the Ceiling}
\addcontentsline{toc}{section}{Teaching 62: Holding from the Bottom Through the Ceiling}

\begin{center}
\jp{第六十二章　底より天井迄持ちの事}
\end{center}

\noindent This teaching concerns the outcome of buying at the bottom,
holding through to the ceiling, and selling before the decline begins.
This is the great way of the market.

This method requires two things: patience and decisiveness. Patience is
the power to hold. Decisiveness is the power to sell.


% =============================================================================
% UPPER VOLUME CHAPTER 63
% =============================================================================
\section*{Teaching 63: A Falling Market as the Starting Point of the Next Rise}
\addcontentsline{toc}{section}{Teaching 63: A Falling Market as the Starting Point of the Next Rise}

\begin{center}
\jp{第六十三章　下相場は次の上げの始の事}
\end{center}

\noindent A falling market connects greatly with the next rising market.
That is, the bottom of a falling market is the starting point of the
next rising market. The end of a falling market is the beginning of a
rising market.

He who knows this principle does not fear a falling market. If one buys
at the bottom of a decline, one reaps the great profits of the coming
rise.


% =============================================================================
% UPPER VOLUME CHAPTER 64
% =============================================================================
\section*{Teaching 64: Rice Prices Following Natural Principles}
\addcontentsline{toc}{section}{Teaching 64: Rice Prices Following Natural Principles}

\begin{center}
\jp{第六十四章　米値は天道の理に従ふの事}
\end{center}

\noindent The price of rice is determined by heaven's fire---that is,
the weather---and by natural principle. It cannot be moved by human
power. Through changes in weather and differences in harvest quality,
the price of rice is naturally determined.

Know this natural principle and follow it without resistance. He who
goes against nature will certainly incur loss.

\bigskip
\textit{Seed:} Rice is a product of heaven and earth. Its price too
follows the principles of heaven and earth. Human speculation may move
prices for a time, but in the end they return to natural principle.


% =============================================================================
% UPPER VOLUME CHAPTER 65
% =============================================================================
\section*{Teaching 65: Abandoning a Position When Rice Reaches Its Extreme Price}
\addcontentsline{toc}{section}{Teaching 65: Abandoning a Position When Rice Reaches Its Extreme Price}

\begin{center}
\jp{第六十五章　米極値にて持を捨つの事}
\end{center}

\noindent The extreme price of rice means the point where prices have
reached their limit. When the extreme price is reached, one should
abandon the position---that is, close it and withdraw. If one
continues to hold past the extreme, profits will diminish and turn to
loss.

To identify the extreme, as described in the preceding chapter, observe
daily highs and lows and perceive changes in crowd sentiment.

\bigskip
\textit{Seed:} One should observe the market with greed removed. Where
there is greed, one will misjudge the extreme. With a calm heart, judge
in accordance with reason.


% =============================================================================
% UPPER VOLUME CHAPTER 66
% =============================================================================
\section*{Teaching 66: Deciding the Selling Price at the Time of Purchase}
\addcontentsline{toc}{section}{Teaching 66: Deciding the Selling Price at the Time of Purchase}

\begin{center}
\jp{第六十六章　買ひの時に賣値を極むの事}
\end{center}

\noindent Bought rice must eventually be sold. At the time of buying,
one must always think about the time of selling. Before buying,
determine at what price you will buy and at what price you will sell.

The ``apportioning of price'' (\jp{分價}) means how to divide the
profit. Take one-third as your profit and yield the remaining
two-thirds to others---this should be your guiding principle.

\bigskip
\textit{Seed:} He who has already determined the selling price at the
time of buying will not waver in his heart. He who buys without
determining when to sell will, even if he gains profit, miss the time
to exit. Determine the exit before the entry. This is the iron rule
of the market.


% =============================================================================
% UPPER VOLUME CHAPTER 67
% =============================================================================
\section*{Teaching 67: Reading the Balance of Buying and Selling Power}
\addcontentsline{toc}{section}{Teaching 67: Reading the Balance of Buying and Selling Power}

\begin{center}
\jp{第六十七章　一市中の賣買の勢ひ讀みの事}
\end{center}

\noindent To read the momentum of buying and selling within a single
session means to compare the strength of buyers and sellers in one
market. When the power of buying is strong, prices rise; when the power
of selling is strong, prices fall. To discern the relative strength of
these forces is essential.


% =============================================================================
% UPPER VOLUME CHAPTER 68
% =============================================================================
\section*{Teaching 68: Not Buying a Market That Fails to Recover}
\addcontentsline{toc}{section}{Teaching 68: Not Buying a Market That Fails to Recover}

\begin{center}
\jp{第六十八章　戻らぬ下げにて買ふべからずの事}
\end{center}

\noindent When a market fails to recover and keeps falling through
entirely, the overall trend is downward. Do not buy on a small bounce.
When prices have fallen through completely, wait for the major bottom.

To buy when there is no sign of recovery is like catching a falling
blade with your bare hand. Identify the major bottom, and only then
buy.


% =============================================================================
% UPPER VOLUME CHAPTER 69
% =============================================================================
\section*{Teaching 69: The Three Pivotal Months and Turning Points for Upward Moves}
\addcontentsline{toc}{section}{Teaching 69: The Three Pivotal Months and Turning Points for Upward Moves}

\begin{center}
\jp{第六十九章　三位の月上り轉換の事}
\end{center}

\noindent When the three pivotal months reach their extreme and the
market begins to rise, this is what is meant by ``reaching upward at
the extreme of the three positions.'' The three positions are January,
May, and September. When the market reaches its extreme in one of these
three months and begins to rise, one should buy boldly.

The extreme of the three pivotal months is a major turning point of the
market. He who identifies this turning point and acts upon it will gain
great profit. This principle is also discussed in Chapter~12.


% =============================================================================
% UPPER VOLUME CHAPTER 70
% =============================================================================
\section*{Teaching 70: Not Buying Large Quantities at High Prices in a Single Region}
\addcontentsline{toc}{section}{Teaching 70: Not Buying Large Quantities at High Prices in a Single Region}

\begin{center}
\jp{第七十章　一國の高値に大量買ふべからずの事}
\end{center}

\noindent Rice bought at prices that have moved against you---that is,
rice bought at high prices---requires great caution. To buy a large
quantity of one region's rice is something one must be very careful
about. If one buys in large quantity at high prices, even a small
decline will produce a large loss.

\bigskip
\textit{Seed:} ``Buying one region'' means buying up all the rice of a
single province. This requires enormous capital. Moreover, if one
corners a region's rice, one invites the resentment of the people. It
goes against the natural order, and in the end one will incur loss.

One must not buy beyond one's strength. To know one's limits is
essential.

Rice bought at high prices turns to loss immediately upon a decline.
But rice bought at low prices still shows profit even after a small
decline. Therefore the price at which one buys is of the greatest
importance---buy at low prices.

Furthermore, when buying in large quantity, do not buy all at once.
Buy in parts, lowering the average. This method is as described in
the preceding chapters.

The purpose of this chapter is to warn against buying large quantities
at high prices. Engrave it well upon your heart.


% =============================================================================
% UPPER VOLUME CHAPTER 71
% =============================================================================
\section*{Teaching 71: Monthly Settlement Cycles and Buying Opportunities}
\addcontentsline{toc}{section}{Teaching 71: Monthly Settlement Cycles and Buying Opportunities}

\begin{center}
\jp{第七十一章　月々の節と大買の機の事}
\end{center}

\noindent The ``monthly settlement pull'' means pulling out at the
settlement points that occur every few months. ``Large buying in the
off-month'' means buying heavily in a month different from the
settlement month. ``Breaking away'' means the market breaking out of a
range.

Know the settlements that occur every few months, and trade at those
junctures. Exercise caution with large buying in off-months. When the
market breaks away, act boldly.


% =============================================================================
% UPPER VOLUME CHAPTER 72
% =============================================================================
\section*{Teaching 72: Year-Long Trading Requires Physical and Mental Health}
\addcontentsline{toc}{section}{Teaching 72: Year-Long Trading Requires Physical and Mental Health}

\begin{center}
\jp{第七十二章　通年の相場は心身の健の事}
\end{center}

\bigskip
\textit{Seed:} ``Year-long trading'' means trading throughout the
entire year. ``Body and mind'' means keeping both body and mind in
order. To trade year-round, both body and mind must be sound.

When the body weakens, judgment weakens, and the heart weakens too.
Keep the body healthy, keep the heart right, and only then approach the
market.

\bigskip
\textbf{(1)} Year-long trading is a contest lasting more than a year, so
soundness of body and mind is of great importance. When fatigued, rest.
When the heart is disordered, be still.


% =============================================================================
% UPPER VOLUME CHAPTER 73
% =============================================================================
\section*{Teaching 73: Observing Three Days Before Committing Large Capital}
\addcontentsline{toc}{section}{Teaching 73: Observing Three Days Before Committing Large Capital}

\begin{center}
\jp{第七十三章　大金を入るには三日を見ての事}
\end{center}

\noindent When committing large capital, observe three days of price
action before moving. See the first day's price, see the second day's
price, see the third day's price, and determine the direction from the
three days' movement.

If prices rise for three days, buy. If prices fall for three days, sell.
The three-day movement reveals the short-term trend.

Committing large capital is a matter of great importance in the market.
Do not act in haste. See the three days' prices, and only after the
direction is established should you act. Moreover, even after committing
large capital, pay attention to each day's prices, and when the momentum
changes, close your position immediately.

Thus the way of the market depends upon patience and decisiveness, the
cultivation of the heart, and the investigation of principle. Savor the
teachings of this book well, and thereby master the way of the market.

\bigskip
\textit{Seed:} The market is the way of man. He who knows the way of
man knows the way of the market.

\begin{editorialnote}
With this chapter, the upper volume concludes. The way of the market
rests upon patience and decisiveness, the cultivation of the mind, and
the investigation of principle. Study the teachings of this book
thoroughly, accumulate practice, and only then approach the market.
This is both the beginning and the end of the way of the market.
\end{editorialnote}

% =============================================================================
\part{Lower Volume \jp{（下巻）}}

\begin{editorialnote}
Both the upper volume (Teachings 1--73) and the lower volume
(Teachings 1--66) that follow have been translated from the
character-level transcription. The lower volume's chapters are
generally shorter than the upper volume's. Many characters in the
transcription carry significant uncertainty, marked with~(?). We
translate the most likely reading in each case. Sections marked
\jp{（一）} introduce the main teaching; sections marked \jp{（種）}
(``Seed'') provide supplementary elaboration.
\end{editorialnote}

\setcounter{chapter}{0}

% =============================================================================
% LOWER VOLUME CHAPTER 1
% =============================================================================
\chapter{Timing and Calendar}
\markboth{Lower Volume: Timing and Calendar}{}

\section*{Teaching 1: The Timing of Declines}
\addcontentsline{toc}{section}{Teaching 1: The Timing of Declines}

\begin{center}
\jp{下巻第一章　落節の事}
\end{center}

\noindent Declines have their own rhythm and timing
(\jp{節}, \textit{setsu}). Within each month there are one or two
regular points of decline. Know these timing points and use them to
determine when to buy and sell. The 15th and month-end are common
inflection points.


% =============================================================================
% LOWER VOLUME CHAPTER 2
% =============================================================================
\section*{Teaching 2: A Three-Month Consolidation and Rotating Capital}
\addcontentsline{toc}{section}{Teaching 2: A Three-Month Consolidation and Rotating Capital}

\begin{center}
\jp{下巻第二章　相場三ヶ月持合金繰り回の事}
\end{center}

\noindent If the market does not move for three months, exit
and rotate your capital elsewhere. Idle capital earns nothing. During
consolidation, deploy funds to other opportunities. When the market
begins to move again, re-enter.


% =============================================================================
% LOWER VOLUME CHAPTER 3
% =============================================================================
\section*{Teaching 3: Trading by Principle, Not by Gambling}
\addcontentsline{toc}{section}{Teaching 3: Trading by Principle, Not by Gambling}

\begin{center}
\jp{下巻第三章}
\end{center}

\noindent Trade by reason (\jt{道理}{d\=ori}), not by luck.
Trading without reason is mere gambling. He who trades by reason profits
long-term; he who trades without reason profits only temporarily and
ultimately loses.


% =============================================================================
% LOWER VOLUME CHAPTER 4
% =============================================================================
\section*{Teaching 4: Entering Positions at Low Prices}
\addcontentsline{toc}{section}{Teaching 4: Entering Positions at Low Prices}

\begin{center}
\jp{下巻第四章　當買を低下に仕掛の事}
\end{center}

\noindent The entry price matters most. Buy at low
prices---even a small rise yields profit. Buy at high prices and only a
large rise yields anything. Use previous bottom prices as reference
points for low-price entries.


% =============================================================================
% LOWER VOLUME CHAPTER 5
% =============================================================================
\section*{Teaching 5: The Three-Year Harvest Cycle}
\addcontentsline{toc}{section}{Teaching 5: The Three-Year Harvest Cycle}

\begin{center}
\jp{下巻第五章　三年結構の方へ國を米達の事}
\end{center}

\noindent Three consecutive years of good harvests lead to
an oversupply of rice and falling prices. Three years of poor harvests
lead to scarcity and rising prices. Know this three-year cycle to read
the overall trend. After abundance comes scarcity; after scarcity comes
abundance---a perpetual cycle.


% =============================================================================
% LOWER VOLUME CHAPTER 6
% =============================================================================
\section*{Teaching 6: The Great Rising Day}
\addcontentsline{toc}{section}{Teaching 6: The Great Rising Day}

\begin{center}
\jp{下巻第六章　天一大上日の事}
\end{center}

\noindent Certain days mark great turning points in the
market. These are known through the calendar (\jt{暦}{koyomi}). When
the market moves on such a day, pay close attention. These major-move
days come only a few times per year---wait for them and act decisively.


% =============================================================================
% LOWER VOLUME CHAPTER 7
% =============================================================================
\section*{Teaching 7: Not Trading Without Adequate Capital}
\addcontentsline{toc}{section}{Teaching 7: Not Trading Without Adequate Capital}

\begin{center}
\jp{下巻第七章　金に際なくして思ふ時合ぶない事}
\end{center}

\noindent Do not trade when your capital is insufficient. To
borrow money to trade or to trade under financial stress impairs
judgment. Trade only with surplus capital. Trading without capital is
like going to battle on an empty stomach---you will have no strength and
will lose.


% =============================================================================
% LOWER VOLUME CHAPTER 8
% =============================================================================
\section*{Teaching 8: The Cyclical Nature of Market Rises and Falls}
\addcontentsline{toc}{section}{Teaching 8: The Cyclical Nature of Market Rises and Falls}

\begin{center}
\jp{下巻第八章　上下の波の如き事}
\end{center}

\noindent ``Rises and falls recurring'' means that a market which has risen will fall again, and a market which has fallen will rise again. The market perpetually repeats its rises and falls.

\bigskip
\noindent ``Reaching one's aim'' means that a position one has initiated reaches its objective. Within the repeated rises and falls, it is essential to reach the objective of one's trade.

\bigskip
\textit{Seed:} The recurring rises and falls are like waves. Know the movement of the waves and ride them to take profit. He who opposes the waves will suffer losses.


% =============================================================================
% LOWER VOLUME CHAPTER 9
% =============================================================================
\section*{Teaching 9: The Laws Governing Major Market Moves}
\addcontentsline{toc}{section}{Teaching 9: The Laws Governing Major Market Moves}

\begin{center}
\jp{下巻第九章　大相場の法の事}
\end{center}

\noindent A ``major market move'' refers to a market of great fluctuation. The movement of major markets has its own laws of momentum. A market that has risen greatly will fall greatly. A market that has fallen greatly will rise greatly. Knowing this law is essential.

\bigskip
\textit{Seed:} The momentum of a major market can extend over several months. Determine the direction of the overall trend and ride its momentum. He who opposes the overall trend will inevitably suffer losses.


% =============================================================================
% LOWER VOLUME CHAPTER 10
% =============================================================================
\section*{Teaching 10: Not Judging the Trend from a Single Day}
\addcontentsline{toc}{section}{Teaching 10: Not Judging the Trend from a Single Day}

\begin{center}
\jp{下巻第十章　一日の動きにて大勢判ずべからずの事}
\end{center}

\noindent A single day's price represents one day's market value. One must not judge the overall trend from a single day's movement. A single day's movement is but a small wave and does not indicate the overall trend.

\noindent Do not let your heart be swayed by a single day's price movement. Observe the movement over several days and several months, and determine the overall trend from that. Rejoicing and despairing over each day's rise and fall is not the way of the market.

\bigskip
\textit{Seed:} To trade based on a single day's movement is no different from gambling. Judge the overall trend by reason, and only then act.


% =============================================================================
% LOWER VOLUME CHAPTER 11
% =============================================================================
\section*{Teaching 11: Trading Twenty-Six Bales and Proper Position Sizing}
\addcontentsline{toc}{section}{Teaching 11: Trading Twenty-Six Bales and Proper Position Sizing}

\begin{center}
\jp{下巻第十一章　二十六俵相場張の事}
\end{center}

\textit{Seed:} Twenty-six bales (\jp{俵}) refers to a quantity of rice, denoting a trade of twenty-six bales. This chapter describes the principles one should know when trading in such quantities.

\noindent The principles of trading differ according to whether the quantity is large or small. For large-quantity trades, one should act with a broad mind. For small-quantity trades, one should act with caution.


% =============================================================================
% LOWER VOLUME CHAPTER 12
% =============================================================================
\section*{Teaching 12: Smaller Quantities and Careful Management}
\addcontentsline{toc}{section}{Teaching 12: Smaller Quantities and Careful Management}

\begin{center}
\jp{下巻第十二章　十七俵少量の事}
\end{center}

\noindent Seventeen bales is a small-lot trade. Even in small-lot trades, the principles are the same as for large lots. Determine the overall trend and trade in accordance with reason.

\noindent Do not be careless simply because the quantity is small. Through the accumulation of small lots, large profits are made.

\bigskip
\textit{Seed:} Whether in large lots or small lots, the way of the market is the same. Act in accordance with reason.

\noindent After gaining profit from seventeen-bale trades, gradually increase your quantity. However, do not trade in quantities exceeding your own strength. Determine the quantity according to your means.

\noindent Furthermore, even in small-lot trades, keep a ledger. Review the ledger to know the results of your own trading. Compare the times of large profit with the times of large loss, and know your own weaknesses.


% =============================================================================
% LOWER VOLUME CHAPTER 13
% =============================================================================
\chapter{Specific Trading Rules}
\markboth{Lower Volume: Specific Trading Rules}{}

\section*{Teaching 13: Trading with One Kan of Capital}
\addcontentsline{toc}{section}{Teaching 13: Trading with One Kan of Capital}

\begin{center}
\jp{下巻第十三章　一貫金の張方の事}
\end{center}

\noindent ``One \textit{kan} purchase'' means buying with one \textit{kan} (\jp{貫目}) of gold. This chapter describes how one should trade when using one \textit{kan} of capital.

\bigskip
\textit{Seed:} When trading with one \textit{kan} of capital, first commit one-third of it to initiate the position, and keep the remainder as reserves. Do not invest the full amount at once.


% =============================================================================
% LOWER VOLUME CHAPTER 14
% =============================================================================
\section*{Teaching 14: The Golden Ratio of Effort and Profit}
\addcontentsline{toc}{section}{Teaching 14: The Golden Ratio of Effort and Profit}

\begin{center}
\jp{下巻第十四章　八分の力三分の利の事}
\end{center}

\noindent ``Eight parts effort, three parts profit'' means using eight-tenths of your strength to buy and taking three-tenths profit as sufficient. One must not use the full ten-tenths of one's strength. Use eight and keep two in reserve.

\noindent Be content with three-tenths profit. If you try to take the full ten-tenths, you may instead lose your gains entirely. This is as stated in Teaching~14 of the upper volume.

\bigskip
\textit{Seed:} Eight-tenths effort for three-tenths profit---this is the golden rule of the market. Keep strength in reserve and restrain your profit-taking. He who observes this principle will sustain himself in the market for a long time.


% =============================================================================
% LOWER VOLUME CHAPTER 15
% =============================================================================
\section*{Teaching 15: Not Buying in Abundant Years}
\addcontentsline{toc}{section}{Teaching 15: Not Buying in Abundant Years}

\begin{center}
\jp{下巻第十五章　豊年に買ふべからずの事}
\end{center}

\noindent Do not buy in abundant years. In abundant years, the supply of rice is plentiful and prices normally fall. If you accumulate buying positions in an abundant year, it will be difficult to endure the decline in prices.

\noindent Therefore, in abundant years, refrain from buying and look for opportunities to sell. Sell at the high prices of an abundant year and wait for lower prices before buying later. This is the way to handle the market in abundant years.

\bigskip
\textit{Seed:} Even in abundant years, prices may rise temporarily. To ride such a temporary rise and buy is dangerous. Since it is only a temporary rise within an overall decline, prices will soon fall again.


% =============================================================================
% LOWER VOLUME CHAPTER 16
% =============================================================================
\section*{Teaching 16: Not Selling Based Solely on Past Patterns}
\addcontentsline{toc}{section}{Teaching 16: Not Selling Based Solely on Past Patterns}

\begin{center}
\jp{下巻第十六章　因果のみにて賣るべからずの事}
\end{center}

\noindent Do not sell by comparing with past results (\jt{因果}{inga}). ``Past results'' means judging based on what happened before. To think that because prices fell in the past they must fall again, and to sell on that basis, is not necessarily correct.

\noindent The past is past, and the present is the present. Look at today's overall trend and make today's judgment. To be captive to the past and lose sight of the present is an error in the way of the market.

\bigskip
\textit{Seed:} To trade based on past results is like walking a road while looking behind you. One should look ahead and walk the road. That is to say, look at today's overall trend and trade accordingly.

\noindent The key insight regarding past results is that one should look at the overall trend and judge the present from it. Do not be captive to the past; follow today's overall trend. When the overall trend is rising, buy without relying on past results. When the overall trend is falling, sell without relying on past results.


% =============================================================================
% LOWER VOLUME CHAPTER 17
% =============================================================================
\section*{Teaching 17: The Halfway Point After a Sharp Decline}
\addcontentsline{toc}{section}{Teaching 17: The Halfway Point After a Sharp Decline}

\begin{center}
\jp{下巻第十七章　天井より急落半値の事}
\end{center}

\noindent A market that falls sharply from the ceiling will decline rapidly to the halfway point (\jt{半値}{hanne}). From the ceiling price to the bottom price, the midpoint---that is, the half-value---is where momentum is strong and the decline is swift.

\bigskip
\textit{Seed:} During a sharp decline, do not buy until the halfway point is reached. After reaching the halfway point, observe whether the bottom is firm, and only then buy. To buy in the middle of a sharp decline is like catching a falling blade with your bare hand.


% =============================================================================
% LOWER VOLUME CHAPTER 18
% =============================================================================
\section*{Teaching 18: Ceilings and Bottoms in Abundant and Famine Years}
\addcontentsline{toc}{section}{Teaching 18: Ceilings and Bottoms in Abundant and Famine Years}

\begin{center}
\jp{下巻第十八章　豊年凶年の天井底の事}
\end{center}

\noindent The way of making the market and the momentum of the trend in abundant years differs from that in famine years. In abundant years, the supply of rice is plentiful and prices naturally fall. At such times, one should take the path of selling.

\noindent In abundant years, the ceiling is low and the bottom deep. In famine years, the ceiling is high and the bottom shallow. Know this difference and determine your method of trading accordingly.

\bigskip
\textit{Seed:} In abundant years, do not try to profit greatly. It is best to take small gains steadily. Famine years are the opportunity to take large profits.


% =============================================================================
% LOWER VOLUME CHAPTER 19
% =============================================================================
\section*{Teaching 19: Do Not Chase Profits That Have Fled}
\addcontentsline{toc}{section}{Teaching 19: Do Not Chase Profits That Have Fled}

\begin{center}
\jp{下巻第十九章　逃げ足を追ふべからずの事}
\end{center}

\noindent Do not chase profits that flee like running feet. Profits depart swiftly, like feet. In chasing profits, you may instead invite losses. Take profits when they come. Do not pursue them after they have gone.


% =============================================================================
% LOWER VOLUME CHAPTER 20
% =============================================================================
\section*{Teaching 20: Not Following the Crowd}
\addcontentsline{toc}{section}{Teaching 20: Not Following the Crowd}

\begin{center}
\jp{下巻第二十章　人に附かぬの事}
\end{center}

\textit{Seed:} It happens that one follows what others do and does the same. When others buy, you buy too; when others sell, you sell too. This is moving in the wake of others.

\noindent However, he who follows others is always late. If you move after others, your profits will be small. Move ahead of others by your own judgment.


% =============================================================================
% LOWER VOLUME CHAPTER 21
% =============================================================================
\section*{Teaching 21: Not Wagering Your Entire Fortune}
\addcontentsline{toc}{section}{Teaching 21: Not Wagering Your Entire Fortune}

\begin{center}
\jp{下巻第廿一章　有金殘らず張るべからずの事}
\end{center}

\noindent Do not stake everything---that is, do not commit your entire resources to trading. He who commits all his resources to the market will be ruined by a single failure. Always keep reserves on hand.


% =============================================================================
% LOWER VOLUME CHAPTER 22
% =============================================================================
\section*{Teaching 22: Three Tests and Seven Days of Confirmation}
\addcontentsline{toc}{section}{Teaching 22: Three Tests and Seven Days of Confirmation}

\begin{center}
\jp{下巻第廿二章　三度の試しと七日の勢ひの事}
\end{center}

\noindent Ceilings and bottoms are settled after being tested three times. Do not determine a ceiling or bottom from a single test. Test three times and then judge.

\noindent When buying, observe the movement over seven days before acting. If prices rise for seven days, the momentum is strong. If prices fall for seven days, the momentum is weak. Use the count of seven days to judge the overall trend.


% =============================================================================
% LOWER VOLUME CHAPTER 23
% =============================================================================
\section*{Teaching 23: The Three Tiers of the Market}
\addcontentsline{toc}{section}{Teaching 23: The Three Tiers of the Market}

\begin{center}
\jp{下巻第廿三章　上中下三段の事}
\end{center}

\noindent The ``three-tier market'' means viewing the market in three stages. The upper stage is near the ceiling, the middle stage is at the half-value, and the lower stage is near the bottom.

\noindent Buy at the lower stage and sell at the upper stage. This is the great road of the market. At the middle stage, since prices may move in either direction, one should refrain from acting.


% =============================================================================
% LOWER VOLUME CHAPTER 24
% =============================================================================
\section*{Teaching 24: July as a Turning-Point Month}
\addcontentsline{toc}{section}{Teaching 24: July as a Turning-Point Month}

\begin{center}
\jp{下巻第廿四章　七月轉換の事}
\end{center}

\noindent July is a month when the market turns. The summer decline may find its bottom in July, and prices may begin rising toward autumn.

\noindent The bottom is reached in July because the outlook for the new rice crop becomes clear. In years of abundant harvest, prices may fall still further from the July bottom. In years of poor harvest, prices may rise greatly from the July bottom.

\noindent The movement in July determines the overall trend for the second half of the year. Pay attention to the July market and set your strategy for autumn and beyond accordingly.

\noindent Furthermore, July follows the sixth-month settlement period, and the consolidation that lasted through June may break. Observe the direction of this breakout and trade accordingly.


% =============================================================================
% LOWER VOLUME CHAPTER 25
% =============================================================================
\section*{Teaching 25: Monitoring the Osaka Market}
\addcontentsline{toc}{section}{Teaching 25: Monitoring the Osaka Market}

\begin{center}
\jp{下巻第廿五章　上方の相場を見るの事}
\end{center}

\noindent The ``upper market'' (\jt{上方}{kamigata}) refers to the \=Osaka market. Know where the \=Osaka market's movement reaches, and use it to judge the Sakata market. The \=Osaka market may move first, and the Sakata market may follow afterward.

\bigskip
\textit{Seed:} It is essential to know the price difference between the \=Osaka and Sakata markets. When the difference is large, it is an easy time to take profits.


% =============================================================================
% LOWER VOLUME CHAPTER 26
% =============================================================================
\section*{Teaching 26: Month-End Settlement Days}
\addcontentsline{toc}{section}{Teaching 26: Month-End Settlement Days}

\begin{center}
\jp{下巻第廿六章　月末の節日の事}
\end{center}

\noindent The ``month-end settlement market'' refers to a market that is settled at the end of the month. The month-end settlement market tends to move as the end of the month approaches.

\bigskip
\textit{Seed:} Pay attention to settlement days. From three days before month-end, settlement trading increases and prices may move.


% =============================================================================
% LOWER VOLUME CHAPTER 27
% =============================================================================
\section*{Teaching 27: Comparing October and August Prices}
\addcontentsline{toc}{section}{Teaching 27: Comparing October and August Prices}

\begin{center}
\jp{下巻第廿七章　十月と八月の値比べの事}
\end{center}

\noindent It is essential to compare the markets in October and August and know the state of markets in all regions. October is the month after the autumn harvest, when the supply of rice is settled. August is before the harvest, when the outlook is still uncertain.

\noindent Compare the prices of these two months and judge the year's harvest quality accordingly.


% =============================================================================
% LOWER VOLUME CHAPTER 28
% =============================================================================
\chapter{The Trader's Calendar}
\markboth{Lower Volume: The Trader's Calendar}{}

\section*{Teaching 28: Unfavorable Calendar Days for Trading}
\addcontentsline{toc}{section}{Teaching 28: Unfavorable Calendar Days for Trading}

\begin{center}
\jp{下巻第廿八章　甲乙の不利日の事}
\end{center}

\noindent Days of the \textit{k\=o-otsu} (\jp{甲乙}) cycle are days when the market does not move. On \textit{k\=o-otsu} days, do not initiate large positions. If you initiate trades on days when the market does not move, commissions pile up with no profit.

\noindent The market has days that move and days that do not move. Choose the days that move and initiate trades on those days.


% =============================================================================
% LOWER VOLUME CHAPTER 29
% =============================================================================
\section*{Teaching 29: Significant Dates for Rice Delivery}
\addcontentsline{toc}{section}{Teaching 29: Significant Dates for Rice Delivery}

\begin{center}
\jp{下巻第廿九章　三月六月の節日の事}
\end{center}

\noindent The sixth and third months have many significant days for rice deliveries. ``Delivery days'' are the days when rice shipments arrive.


% =============================================================================
% LOWER VOLUME CHAPTER 30
% =============================================================================
\section*{Teaching 30: The Fifth and Seventh Day as Inflection Points}
\addcontentsline{toc}{section}{Teaching 30: The Fifth and Seventh Day as Inflection Points}

\begin{center}
\jp{下巻第三十章　五日目七日目の節の事}
\end{center}

\noindent The ``gates of five and seven'' mean that inflection points occur at the fifth and seventh day of a market move. If prices rise for five days, they will pull back once; if they rise for seven days, they will pull back again. Knowing these gates of five and seven is essential.


% =============================================================================
% LOWER VOLUME CHAPTER 31
% =============================================================================
\section*{Teaching 31: Autumn Is the Time to Buy}
\addcontentsline{toc}{section}{Teaching 31: Autumn Is the Time to Buy}

\begin{center}
\jp{下巻第三十一章　秋は買ひの時の事}
\end{center}

\noindent Autumn is the season of the rice harvest, and prices are normally low. Accumulate buying positions at this cheap time. Buy at autumn's low prices and sell at winter's high prices. This is the standard way of the market.


% =============================================================================
% LOWER VOLUME CHAPTER 32
% =============================================================================
\section*{Teaching 32: Tracking Daily Momentum}
\addcontentsline{toc}{section}{Teaching 32: Tracking Daily Momentum}

\begin{center}
\jp{下巻第三十二章　日勢の事}
\end{center}

\noindent Daily momentum (\jt{日勢}{nissei}) means the market's momentum each day. Observe the daily momentum and use it to judge the overall trend. Just as weapons and tools are useful only when polished every day, so too with the market. Knowing daily momentum is the foundation of trading.


% =============================================================================
% LOWER VOLUME CHAPTER 33
% =============================================================================
\section*{Teaching 33: Unlucky Days When the Market Is Prone to Sharp Moves}
\addcontentsline{toc}{section}{Teaching 33: Unlucky Days When the Market Is Prone to Sharp Moves}

\begin{center}
\jp{下巻第三十三章　厄日の事}
\end{center}

\noindent Unlucky days (\jt{厄日}{yakubi}) are days when the market is prone to sharp fluctuation. On unlucky days, there can be large moves. Pay attention to these days and act when you see the opportunity. Also, since great losses may occur on unlucky days, exercise caution.


% =============================================================================
% LOWER VOLUME CHAPTER 34
% =============================================================================
\section*{Teaching 34: Quarter-End Settlement Points}
\addcontentsline{toc}{section}{Teaching 34: Quarter-End Settlement Points}

\begin{center}
\jp{下巻第三十四章　四半期の節の事}
\end{center}

\textit{Seed:} The end of March, the end of June, the end of September, and the end of December are critically important seasonal junctures. At these month-ends, settlement trading is heavy and the market may move sharply.


% =============================================================================
% LOWER VOLUME CHAPTER 35
% =============================================================================
\section*{Teaching 35: Buying in Three Installments Around Settlement Days}
\addcontentsline{toc}{section}{Teaching 35: Buying in Three Installments Around Settlement Days}

\begin{center}
\jp{下巻第三十五章　節日の前中後三度買ひの事}
\end{center}

\noindent There is a method of buying in three large installments around settlement days. Buy in three parts: before, on, and after the settlement day. The first purchase is before the settlement day, the second on the day itself, and the third after the settlement day. By dividing into three purchases in this manner, one can prepare for the movements around the settlement.


% =============================================================================
% LOWER VOLUME CHAPTER 36
% =============================================================================
\section*{Teaching 36: Identifying the Day the Market's Direction Is Established}
\addcontentsline{toc}{section}{Teaching 36: Identifying the Day the Market's Direction Is Established}

\begin{center}
\jp{下巻第三十六章　立て局の日の事}
\end{center}

\noindent The ``day the direction is established'' is the day when the market's overall trend becomes settled. On this day, there can be a large move. Identify the day the direction is established and act without letting the opportunity escape.


% =============================================================================
% LOWER VOLUME CHAPTER 37
% =============================================================================
\section*{Teaching 37: Reading the Tail of a Move}
\addcontentsline{toc}{section}{Teaching 37: Reading the Tail of a Move}

\begin{center}
\jp{下巻第三十七章　勢ひの尾を讀むの事}
\end{center}

\noindent The ``tail'' means the tail-end of the market's momentum. The ``two tails'' are the tail of the rise and the tail of the decline. The tail of the rise is near the ceiling, where momentum wanes. The tail of the decline is near the bottom, where momentum is exhausted.

\noindent Discerning these two tails is a matter of great importance in the way of the market.


% =============================================================================
% LOWER VOLUME CHAPTER 38
% =============================================================================
\section*{Teaching 38: Major Moves Begin After Settlement Days}
\addcontentsline{toc}{section}{Teaching 38: Major Moves Begin After Settlement Days}

\begin{center}
\jp{下巻第三十八章　節後の大動きの事}
\end{center}

\textit{Seed:} That ``great moves begin after settlement days'' means that after a settlement day, the market may begin to move greatly. Since settlement days are turning points in the market, pay close attention to the movement after settlement days.


% =============================================================================
% LOWER VOLUME CHAPTER 39
% =============================================================================
\section*{Teaching 39: The Five Rules of the Rice Merchant's Mindset}
\addcontentsline{toc}{section}{Teaching 39: The Five Rules of the Rice Merchant's Mindset}

\begin{center}
\jp{下巻第三十九章　米商の心得五則の事}
\end{center}

\noindent The rice merchant's mental discipline is the root of the way of the market. The rice merchant should always keep his heart right and act in accordance with reason. He whose heart is right will gain profit; he whose heart is crooked will suffer loss.

\noindent Observe the overall trend, know the seasons, perceive market sentiment, and only then trade. When the heart is unsettled, rest. When the heart is composed, act.

\bigskip
\textit{Seed:} The rice merchant's principles are as follows:
\begin{itemize}
\item Act in accordance with reason.
\item Abandon greed.
\item Observe the overall trend before moving.
\item Wait with patience.
\item Keep the heart right when engaging the market.
\end{itemize}
He who observes these five rules will sustain himself as a rice merchant.


% =============================================================================
% LOWER VOLUME CHAPTER 40
% =============================================================================
\section*{Teaching 40: Approaching the Market with a Single Great Resolve}
\addcontentsline{toc}{section}{Teaching 40: Approaching the Market with a Single Great Resolve}

\begin{center}
\jp{下巻第四十章　一大決心の事}
\end{center}

\noindent ``One great mental resolve'' means having a single, large way of holding one's mind. When approaching the market, unify your mind and think broadly. Do not be captive to small matters; see things on a grand scale.


% =============================================================================
% LOWER VOLUME CHAPTER 41
% =============================================================================
\section*{Teaching 41: Buying in the Month That Heaven Has Ordained}
\addcontentsline{toc}{section}{Teaching 41: Buying in the Month That Heaven Has Ordained}

\begin{center}
\jp{下巻第四十一章　天の定めし月に買ふの事}
\end{center}

\textit{Seed:} The month to buy is the month when heaven has set the bottom. Buy at the bottom that heaven has ordained. It is not a bottom determined by human judgment, but a bottom determined by heaven.

\noindent The ``bottom set by heaven'' means a bottom determined by natural principles. Through changes in harvest abundance and famine, and through seasonal movements, a natural bottom is reached. Discern this natural bottom and buy.


% =============================================================================
% LOWER VOLUME CHAPTER 42
% =============================================================================
\section*{Teaching 42: Withdrawing and Saving Market Profits}
\addcontentsline{toc}{section}{Teaching 42: Withdrawing and Saving Market Profits}

\begin{center}
\jp{下巻第四十二章　利を引きて貯むの事}
\end{center}

\noindent Profits from winning should be withdrawn and set aside---that is, the money gained from the market should be taken out and saved separately. If money gained from profits is reinvested in the market, the profits will quickly vanish and turn to losses.

\bigskip
\textit{Seed:} When you gain a profit, be sure to save a portion of it separately. Do not reinvest everything in the market. Save profits and hold them as reserves against times of failure. This is the way to continue in the market for a long time.


% =============================================================================
% LOWER VOLUME CHAPTER 43
% =============================================================================
\chapter{Final Teachings}
\markboth{Lower Volume: Final Teachings}{}

\section*{Teaching 43: The Five Forms of Market Movement}
\addcontentsline{toc}{section}{Teaching 43: The Five Forms of Market Movement}

\begin{center}
\jp{下巻第四十三章　五つの相場形の事}
\end{center}

\textit{Seed:} The market should be approached through five forms. The five forms are: rising, falling, consolidation, upward breakout, and downward breakout. Know these five forms and act according to each.


% =============================================================================
% LOWER VOLUME CHAPTER 44
% =============================================================================
\section*{Teaching 44: Reading the Role of Daily Price Movements}
\addcontentsline{toc}{section}{Teaching 44: Reading the Role of Daily Price Movements}

\begin{center}
\jp{下巻第四十四章　日々の値動きの役の事}
\end{center}

\noindent ``Reading the role of daily price movements'' means observing the daily price movements and knowing their role, so as to reach the objective of your trade.


% =============================================================================
% LOWER VOLUME CHAPTER 45
% =============================================================================
\section*{Teaching 45: The Ruin of Trading Beyond One's Means}
\addcontentsline{toc}{section}{Teaching 45: The Ruin of Trading Beyond One's Means}

\begin{center}
\jp{下巻第四十五章　分限を超えて張る滅亡の事}
\end{center}

\noindent This concerns those who invest large sums beyond their means, devote themselves entirely, and bring misfortune upon their lives. That is, those who invest large sums in the market and lose them, ruining themselves.

\noindent Those who lose large sums in the market are those who engage in contests exceeding their own strength. Trade within your means and engage in contests that are not unreasonable.


% =============================================================================
% LOWER VOLUME CHAPTER 46
% =============================================================================
\section*{Teaching 46: Accumulating Capital Gradually}
\addcontentsline{toc}{section}{Teaching 46: Accumulating Capital Gradually}

\begin{center}
\jp{下巻第四十六章　少しづつの積み上げの事}
\end{center}

\noindent One should use money like piling up a mountain, reaching one's goal only afterward. Accumulate capital little by little and build it into a large sum. Do not spend large sums all at once.

\bigskip
\textit{Seed:} In using money, one must observe moderation. He who uses money without moderation will ultimately lose it all.


% =============================================================================
% LOWER VOLUME CHAPTER 47
% =============================================================================
\section*{Teaching 47: Not Blindly Following Famous Traders}
\addcontentsline{toc}{section}{Teaching 47: Not Blindly Following Famous Traders}

\begin{center}
\jp{下巻第四十七章　名人に盲目に附くべからずの事}
\end{center}

\noindent Learning the momentum of famous buyers but failing to notice when the market peaks and falls---this happens when one follows the buying momentum of well-known traders without one's own judgment.

\noindent To watch others buy and then buy, or watch others sell and then sell, is merely following others and is not one's own path. Act on your own judgment.

\bigskip
\textit{Seed:} Following the movements of a famous market trader may appear safe at first glance, but if you do not know the timing of his exit, you may suffer losses.


% =============================================================================
% LOWER VOLUME CHAPTER 48
% =============================================================================
\section*{Teaching 48: Market Momentum Is Self-Sustaining}
\addcontentsline{toc}{section}{Teaching 48: Market Momentum Is Self-Sustaining}

\begin{center}
\jp{下巻第四十八章　勢ひは勢ひを呼ぶの事}
\end{center}

\noindent The market's momentum simply persists. That is, if there is upward momentum, it continues upward as it is; if there is downward momentum, it continues downward as it is. Do not oppose the momentum.

\bigskip
\textit{Seed:} Riding the momentum is a matter of great importance in the way of the market. He who opposes the momentum will suffer losses. He who follows the momentum will gain profit.


% =============================================================================
% LOWER VOLUME CHAPTER 49
% =============================================================================
\section*{Teaching 49: Carrying Positions Through Month-End Settlements}
\addcontentsline{toc}{section}{Teaching 49: Carrying Positions Through Month-End Settlements}

\begin{center}
\jp{下巻第四十九章　月末を越すの事}
\end{center}

\noindent This is the method of buying and carrying positions over the month-end to reach one's objective. Carrying positions over the month-end requires both surplus capital and patience.

\bigskip
\textit{Seed:} Carrying positions over the month should be employed only in years when the overall trend is upward. If you carry positions over in a year when the trend is unsettled, you will idle your capital with no profit.


% =============================================================================
% LOWER VOLUME CHAPTER 50
% =============================================================================
\section*{Teaching 50: The Principle of Water}
\addcontentsline{toc}{section}{Teaching 50: The Principle of Water}

\begin{center}
\jp{下巻第五十章　水の理の事}
\end{center}

\noindent The ``principle of water'' in buying means that, like water seeking a low point, one should buy at low prices. ``Reaching one's objective and pulling up'' means withdrawing after reaching your target. ``The ship's error of judgment'' means aspiring to move like a great ship and making an error of the heart.

\noindent Just as setting out to sea in a small boat is dangerous, so is entering a major market with small capital. Know the size of your own vessel and act accordingly.

\bigskip
\textit{Seed:} The principle of water is the foundation of the market. Water flows from high to low and naturally becomes level. The market is the same. When prices are high, they naturally fall; when prices are low, they naturally rise. Follow this natural principle.

\noindent Furthermore, the ship's error of judgment arises from the greed of wanting to take large profits. Abandon greed and act in accordance with reason.


% =============================================================================
% LOWER VOLUME CHAPTER 51
% =============================================================================
\section*{Teaching 51: Combining Monthly Positions into a Large Holding}
\addcontentsline{toc}{section}{Teaching 51: Combining Monthly Positions into a Large Holding}

\begin{center}
\jp{下巻第五十一章　月々の合せ大張りの事}
\end{center}

\noindent ``Combining monthly positions and drawing momentum'' means accumulating purchases over several months, combining their terms, and building them into a large force of momentum.

\noindent Even if each month's purchase is small, combining the purchases of several months makes it large. This method is a way of gradually increasing one's position and is akin to the method of averaging.

\bigskip
\textit{Seed:} The method of combining monthly positions and drawing momentum should be employed in years when the overall trend is upward. If this method is used when the trend is unsettled, it will only enlarge one's losses.

\noindent Furthermore, when combining monthly positions, one should know the average purchase price for each month. If the market rises above the average price, there is profit; if it falls below, there is loss. It is essential always to know the average price.

\noindent The meaning of this chapter is to describe the method of gradually increasing one's purchases to obtain a large profit. It should be employed only after the overall trend has been ascertained.


% =============================================================================
% LOWER VOLUME CHAPTER 52
% =============================================================================
\section*{Teaching 52: Monitoring Both the Osaka and Sakata Markets}
\addcontentsline{toc}{section}{Teaching 52: Monitoring Both the Osaka and Sakata Markets}

\begin{center}
\jp{下巻第五十二章　上方と酒田の相場見合せの事}
\end{center}

\textit{Seed:} This describes combining the momentum of the \=Osaka market with the reaching momentum of the Sakata market, and planning the deployment of capital accordingly. Observe both the \=Osaka and the Sakata market momentum, and determine how to use your capital.

\noindent When the \=Osaka momentum is strong and the Sakata momentum is also strong, buy boldly. When \=Osaka is weak and Sakata is strong, buy with restraint.


% =============================================================================
% LOWER VOLUME CHAPTER 53
% =============================================================================
\section*{Teaching 53: Keeping One's Heart Right and Following Reason}
\addcontentsline{toc}{section}{Teaching 53: Keeping One's Heart Right and Following Reason}

\begin{center}
\jp{下巻第五十三章　心正しく理に從ふの事}
\end{center}

\noindent The most important thing in the rice market is to keep one's heart right and follow reason. If the heart is crooked, one will not gain profit. If one acts against reason, one will suffer losses.

\noindent The way of the market is the way of humanity. He who observes the way of humanity will observe the way of the market as well. He who departs from the way of humanity will depart from the way of the market as well.

\bigskip
\textit{Seed:} Rice is the foundation of the realm. He who trades in rice is involved in the great affairs of the realm. One must not forget this awareness.


% =============================================================================
% LOWER VOLUME CHAPTER 54
% =============================================================================
\section*{Teaching 54: Reading the Extreme Low Point and Buying Momentum}
\addcontentsline{toc}{section}{Teaching 54: Reading the Extreme Low Point and Buying Momentum}

\begin{center}
\jp{下巻第五十四章　極安と買ひの勢ひの事}
\end{center}

\textit{Seed:} The extreme of low prices is the point where the bottom has been reached. Only those who know the direction of market momentum can buy at the extreme low.

\noindent The extreme bottom is reached when prices test the bottom three times and fail to decline further. This is the time to buy.


% =============================================================================
% LOWER VOLUME CHAPTER 55
% =============================================================================
\section*{Teaching 55: Confirming the Bottom's Firmness Before Buying}
\addcontentsline{toc}{section}{Teaching 55: Confirming the Bottom's Firmness Before Buying}

\begin{center}
\jp{下巻第五十五章　底堅めを見て買ふの事}
\end{center}

\noindent The ``footing'' of a bottom market is the firm ground of the bottom. When the footing is firm, buy with confidence. When the footing is not yet settled, continue to wait.

\bigskip
\textit{Seed:} Determining the footing is settled when the bottom has been tested three times without breaking. This method is as stated in the upper volume.


% =============================================================================
% LOWER VOLUME CHAPTER 56
% =============================================================================
\section*{Teaching 56: Accumulating Many Small Wins}
\addcontentsline{toc}{section}{Teaching 56: Accumulating Many Small Wins}

\begin{center}
\jp{下巻第五十六章　小勝をかさぬの事}
\end{center}

\noindent Count and divide your wins, and establish victories to accumulate numbers. That is, pile up winning trades and accumulate profit. Rather than one large win, it is better to accumulate many small wins.


% =============================================================================
% LOWER VOLUME CHAPTER 57
% =============================================================================
\section*{Teaching 57: Progressively Lower Ceilings Signal a Major Decline}
\addcontentsline{toc}{section}{Teaching 57: Progressively Lower Ceilings Signal a Major Decline}

\begin{center}
\jp{下巻第五十七章　月々の天井下がりの事}
\end{center}

\noindent A market that makes a ceiling every month may be turning to a decline. When each month's ceiling becomes progressively lower, it is a sign that the overall trend is turning downward. Observe this sign and determine the right moment to sell.


% =============================================================================
% LOWER VOLUME CHAPTER 58
% =============================================================================
\section*{Teaching 58: Unlucky Days of the Month}
\addcontentsline{toc}{section}{Teaching 58: Unlucky Days of the Month}

\begin{center}
\jp{下巻第五十八章　月の八の日の事}
\end{center}

\noindent Days with the number eight---the 8th, 18th, and 28th---are unfavorable for buying. These are inauspicious days for the market, and it is not advisable to initiate positions on them.


% =============================================================================
% LOWER VOLUME CHAPTER 59
% =============================================================================
\section*{Teaching 59: Building Positions in Three Layers for Major Opportunities}
\addcontentsline{toc}{section}{Teaching 59: Building Positions in Three Layers for Major Opportunities}

\begin{center}
\jp{下巻第五十九章　三段の構えの事}
\end{center}

\noindent The ``threefold plan'' means positioning in three layers. Extended over three rounds of trading, one aims to achieve a great profit. That is, by initiating positions across three rounds, one gains a large profit.


% =============================================================================
% LOWER VOLUME CHAPTER 60
% =============================================================================
\section*{Teaching 60: Comparing Mid-Month and Month-End Prices in May}
\addcontentsline{toc}{section}{Teaching 60: Comparing Mid-Month and Month-End Prices in May}

\begin{center}
\jp{下巻第六十章　五月半ばと月末の値比べの事}
\end{center}

\noindent Compare mid-May prices with month-end prices. When prices at month-end are lower than at mid-month, they will continue falling in June. When prices at month-end are higher than at mid-month, they may continue rising in June.


% =============================================================================
% LOWER VOLUME CHAPTER 61
% =============================================================================
\section*{Teaching 61: Even a Half-Profit Is a Success}
\addcontentsline{toc}{section}{Teaching 61: Even a Half-Profit Is a Success}

\begin{center}
\jp{下巻第六十一章　半利も首尾の事}
\end{center}

\noindent Even winning once and taking only half the possible profit may be sufficient to reach one's objective. Be content with half the profit. Do not lose your gains by attempting to capture the full amount.


% =============================================================================
% LOWER VOLUME CHAPTER 62
% =============================================================================
\section*{Teaching 62: Buying at the Midpoint of a Seasonal Cycle}
\addcontentsline{toc}{section}{Teaching 62: Buying at the Midpoint of a Seasonal Cycle}

\begin{center}
\jp{下巻第六十二章　季節の半ばに買ふの事}
\end{center}

\noindent ``Buying with a name at the midpoint of the season'' means buying at the midpoint of a seasonal cycle. There is a method of buying at the midpoint without waiting for the seasonal juncture. This method moves ahead of the seasonal juncture, making it easy to gain profit.


% =============================================================================
% LOWER VOLUME CHAPTER 63
% =============================================================================
\section*{Teaching 63: Aligning the Right Month with the Right Mental State}
\addcontentsline{toc}{section}{Teaching 63: Aligning the Right Month with the Right Mental State}

\begin{center}
\jp{下巻第六十三章　買月と心と合はすの事}
\end{center}

\textit{Seed:} This describes the match between the buying month and one's mental state. He who knows the right month to buy and acts with a correct heart will reach his objective. He who mistakes the month to buy, or whose heart is not right, will fail to reach his objective and suffer losses.


% =============================================================================
% LOWER VOLUME CHAPTER 64
% =============================================================================
\section*{Teaching 64: Year-Long Trading and the Way of the Four Seasons}
\addcontentsline{toc}{section}{Teaching 64: Year-Long Trading and the Way of the Four Seasons}

\begin{center}
\jp{下巻第六十四章　通年と四時の道の事}
\end{center}

\noindent To trade ``through the year'' means to conduct market operations throughout the entire year. Within a year, there are times of rise, times of decline, and times of consolidation. Act according to each of these periods.

\bigskip
\textit{Seed:} Year-long trading requires both body and mind to be sound. To keep one's heart right and not err in judgment over the long span of a year is an exceedingly difficult undertaking. Yet he who masters this path will gain great profit.

\bigskip
\textit{Seed:} The principles of year-long trading are as follows:

\textbf{(1)} Spring is the time for buying.

\textbf{(2)} Summer is the time for resting.

\textbf{(3)} Autumn is the time for building positions.

\textbf{(4)} Winter is the time for taking profit.

\noindent Follow these seasonal principles and trade through the year. Do not oppose the movement of the four seasons. Act in accordance with nature.

\noindent With this chapter, the principal teachings of the lower volume are substantially concluded.


% =============================================================================
% LOWER VOLUME CHAPTER 65
% =============================================================================
\section*{Teaching 65: Reading Agricultural Reports and Weather}
\addcontentsline{toc}{section}{Teaching 65: Reading Agricultural Reports and Weather}

\begin{center}
\jp{下巻第六十五章　農報と天氣を讀むの事}
\end{center}

\noindent There are times when one judges the market based on agricultural reports. The success or failure of the harvest is a matter of great importance in determining the overall market trend. Know the agricultural reports and use them to judge whether the harvest is abundant or poor.

\bigskip
\textit{Seed:} Whether the harvest is abundant or poor depends greatly on the weather in June and July. Observe the weather during this period carefully and use it to forecast the autumn harvest.


% =============================================================================
% LOWER VOLUME CHAPTER 66
% =============================================================================
\section*{Teaching 66: The Ultimate Teaching: A Right Heart Prevails}
\addcontentsline{toc}{section}{Teaching 66: The Ultimate Teaching: A Right Heart Prevails}

\begin{center}
\jp{下巻第六十六章　正心にて終に勝つの事}
\end{center}

\noindent Even when one enters with a resolute heart, there are times when the momentum does not reach its mark and one cannot win. Even entering with a correct heart, depending on the momentum of the times, there are occasions when one cannot prevail. Yet he whose heart is right will, through the accumulation of many trades, ultimately prevail.

\bigskip
\textit{Seed:} He whose heart is right is not broken by a temporary defeat. Through the accumulation of many victories, he gains great profit. This is the ultimate meaning of the way of the market, and the conclusion of this book.

\begin{editorialnote}
With this chapter, the lower volume and the complete work conclude. One
page of the source archive (page 88) is missing, marked \jp{欠}
(lacking). It likely contained the formal end-of-volume marker and
possibly additional concluding text.
\end{editorialnote}

% =============================================================================
\chapter*{Appendix: Historical Rice Price Data}
\addcontentsline{toc}{chapter}{Appendix: Historical Rice Price Data}
\markboth{Historical Rice Price Data}{}

The original text concludes with five pages of historical rice price
data from the An'ei era (\jp{安永}, 1772--1781) at the Sakata rice
market. The data is organized by year with monthly prices in
\textit{hy\=o} (\jp{俵}, bales) and \textit{bu} (\jp{分}), and
summary annotations describing each year's major market events.
The running header references Chapter~58 (\jp{第五十八}), indicating
this data was meant as empirical evidence for the teachings.

The following prices were read directly from the source scans. Question
marks indicate uncertain characters in the transcription.

\bigskip

\subsection*{An'ei 5, Year of the Monkey (1776)}
\noindent\small\jp{安永五甲年（申）}\normalsize

\begin{tabular}{@{}lrl@{}}
\toprule
\textbf{Month} & \textbf{Bales} & \textbf{Bu} \\
\midrule
1st month \small\jp{（正月）} & 31 & 3 \\
5th month \small\jp{（五月）} & 26 & 7 \\
6th month \small\jp{（六月）} & 41 & --- \\
5th month, 2nd entry(?) & 38 & --- \\
\bottomrule
\end{tabular}

\medskip
\noindent\textit{Annotations:}
\begin{itemize}[nosep,leftmargin=1em]
\item Ceiling reversal on an unlucky day; 40-bale rise.
\item 43-bale decline.
\end{itemize}

\bigskip

\subsection*{An'ei 6, Year of the Rooster (1777)}
\noindent\small\jp{安永六百年（酉）}\normalsize

\begin{tabular}{@{}lrl@{}}
\toprule
\textbf{Period} & \textbf{Bales} & \\
\midrule
Early in year(?) & 38 & (?) \\
12th month, 2nd day & 24 & (?) \\
3rd month(?) & 27 & (?) \\
\bottomrule
\end{tabular}

\medskip
\noindent\textit{Annotations:}
\begin{itemize}[nosep,leftmargin=1em]
\item 5-day alternating ceiling; 140-bale rise(?).
\end{itemize}

\bigskip

\subsection*{Additional Years (pages 90--93)}

The data continues across four more pages covering approximately six to
eight additional years of the An'ei era. The format is consistent: monthly
prices in bales and \textit{bu}, with year-end annotations describing
peaks (\jp{天井}), declines (\jp{下げ}), and the magnitude and duration
of each move. Prices visible in the later pages include:

\begin{itemize}
\item Values ranging from approximately 13 bales to 41 bales
\item Annotations referencing moves of 30--140 bales
\item Zodiac year markers: \jp{申} (Monkey), \jp{酉} (Rooster),
  \jp{戌} (Dog), \jp{亥} (Boar), \jp{子} (Rat), \jp{丑} (Ox),
  \jp{寅} (Tiger), etc.
\end{itemize}

\begin{editorialnote}
The individual price entries on pages 90--93 are legible in the source
scans but require careful character-by-character reading to transcribe
fully. The two year-sections above (An'ei~5 and~6) represent the
clearest data from the first page of the tables. A complete
transcription of all price data across pages 89--93 would be a
valuable addition to a future edition.

One page of the source archive (page~88) is missing, marked \jp{欠}
(lacking). It likely contained the \jp{下巻了} (end of lower volume)
marker.
\end{editorialnote}

\bigskip

\noindent The concluding note to the price data reads:

\begin{quote}
\small\jp{右の年表は、安永年間の酒田米相場の實際の動きを記したるものなり。}

\smallskip

\small\jp{此の實例を以て、本書の教へを確かめ、以て己が相場の道に活かすべし。}

\medskip

\noindent\textit{The year-tables above record the actual movements of the
Sakata rice market during the An'ei era. Use these real examples to
verify the teachings of this book, and apply them to your own way of
the market.}
\end{quote}

\noindent And a final observation:

\begin{quote}
\small\jp{二十五俵位にては天井投落し、}

\small\jp{二十俵位にては大底を打つ。}

\small\jp{此の如く、實際の數字を以て知るべし。}

\medskip

\noindent\textit{At around twenty-five bales, the ceiling reverses and
falls. At around twenty bales, the major bottom is reached. Know these
things through actual numbers.}
\end{quote}

% =============================================================================
\chapter*{About the Editor}
\addcontentsline{toc}{chapter}{About the Editor}
\markboth{About the Editor}{}

Zach Booth is a software engineer at NASA Ames Research Center, where
he uses AI systems---including Claude (Anthropic)---in his daily
engineering workflow. Self-taught in Japanese since the age of eighteen
and conversational in the language, he identified this text in a
digital archive and recognized that no English translation existed.

He directed the translation using the same AI tools he works with
professionally, separating transcription from translation and
publishing the complete audit trail so that every editorial decision
can be checked. The full pipeline---original scans, character-by-character
transcription with confidence levels, working translation files, and
\LaTeX{} source---is maintained alongside the published text.

This is the second volume in his series of translations from the
Japanese market classics, following \textit{Sanen Kinsen Hiroku}
(Three Monkeys Gold Spring Secret Record), published earlier in 2026.

% =============================================================================
\chapter*{Colophon}
\addcontentsline{toc}{chapter}{Colophon}
\markboth{Colophon}{}

\begin{center}
\jp{宗久翁相場全集　終}\\[1em]
\end{center}

\noindent\textbf{Printed:} April 10, Meiji 43 (1910)\\
\textbf{Published:} April 15, Meiji 43 (1910)\\
\textbf{Author/Editor:} Hayasaka H\=oz\=o (\jp{早坂豐蔵})\\
\textbf{Publisher:} Takei Tar\=o (\jp{武井太郎})\\
\textbf{Publishing House:} Shin'gid\=o Shoten (\jp{信義堂書店}), Tokyo\\[1em]

\noindent\textit{Reproduction prohibited} (\jp{不許複製})---now public domain.

\end{document}
